### 2020-05-14 paper.tex
Initial addition.

\documentclass{article}
\usepackage[round]{natbib}

% local definitions
\newcommand{\msprime}[0]{\texttt{msprime}}
\newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}

\begin{document}

\title{msprime 1.0: an efficient and extensible coalescent simulation framework}
\author{A permutation of (AR, DN, JK, SG)}

\abstract{

}

\section*{Introduction}

The \msprime~\citep{kelleher2016efficient} coalescent simulator is now quite
widely used.

\bibliographystyle{plainnat}
\bibliography{paper}

\end{document}

### 2020-05-14 paper.tex
First pass at outline narrative.

@@
 
+\usepackage{listings}
+
+\lstset{language=Python}
+
 % local definitions
 \newcommand{\msprime}[0]{\texttt{msprime}}
+\newcommand{\ms}[0]{\texttt{ms}}
 \newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
@@
 
-\title{msprime 1.0: an efficient and extensible coalescent simulation framework}
+\title{Describing demographic models is hard}
 \author{A permutation of (AR, DN, JK, SG)}
+\maketitle
 
 \abstract{
+We describe two errors made in defining population genetic models using the
+\msprime\ coalescent simulator.
 
@@
 
-The \msprime~\citep{kelleher2016efficient} coalescent simulator is now quite
-widely used.
+The \msprime\ coalescent simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
+is now quite widely used. The large increases in efficiency over the classical
+\ms\ program~\citep{hudson2002generating} make it feasible to simulate large
+samples of whole chromosomes for the first time. Another distinct advantage
+of \msprime\ is the Python API that is its primary interface, greatly
+increasing the flexibility and efficiency over the standard approach of
+text-based command line interfaces. In particular, programs like \ms\
+required users to specify cryptic command line options to describe demographic
+models [JK: maybe include an example ms command line?]. Particularly for
+the large models we are using today, these are not comprehensible for humans.
+The Python API for \msprime\ is a great improvement, allowing the user to
+state models in a documented and reproducible manner. [JK: maybe show
+the same model as described above in msprime notation?]
+
+Describing these models of population history is still hard and error
+prone, however. In this note we discuss how two poor design decisions
+in \msprime's demography API lead to errors being made, which ended
+up in the scientific record.
+
+\section{A bad tutorial example}
+
+To illustrate the demography API, \msprime\ included a description of the
+three population Out-of-Africa model~\citep{gutenkunst2009inferring}
+as part of its tutorial documentation. This example model, however,
+was not correctly implemented.
+
+[What's wrong with the model, and how does this differ from the original. Some
+analysis.]
+
+This model was subsequently copied
+several times and used in papers [JK: do we want to estimate how
+many times it was copied? A quick search on github suggests around 30
+times].
+
+Arguably, this error occured because of a poor API design choice.
+The MassMigration event is confusing and poorly explained. In
+\msprime\ 1.0 we're introducing a PopulationSplit event, which allows
+such models to be described more declaritively.
+
+\section{A missing parameter}
+
+[Description of what went wrong in the AJHG paper~\citep{martin2017human}.
+Some analysis of how this affects the conclusions of the paper.]
+
+[How was the error found? I wonder if the reason that Alicia didn't
+do any basic checks on the simulation output was because it was
+so massive. Until recently we couldn't do things like compute Fst, etc
+on the tree sequences. Still, good enough reason to run small simulations
+first to make sure that you're doing something sensible.]
+
+This error was clearly caused by poor API design. By forcing the user to
+pass around three separate parameters to describe the demographic model,
+it's inviting simple mistakes such as this. In \msprime\ 1.0 we introduce
+a Demography class, which wraps these three parameters.
+Thus, instread of writing
+\begin{lstlisting}[frame=single]
+dbg = msprime.DemographyDebugger(
+    population_configurations=population_configurations,
+    migration_matrix=migration_matrix,
+    demographic_events=demographic_events)
+dbg.print_history()
+ts = msprime.simulate(
+    population_configurations=population_configurations,
+    migration_matrix=migration_matrix,
+    demographic_events=demographic_events)
+\end{lstlisting}
+we would now write
+\begin{lstlisting}[frame=single]
+dbg = demography.debug()
+dbg.print_history()
+ts = msprime.simulate(demography=demography)
+\end{lstlisting}
+
+\section{Conclusions}
+
+Stuff to mention in no particular order:
+\begin{itemize}
+\item Defining demographic models is hard, and we need all the help we can get.
+   We really need \stdpopsim.
+\item If you do have to implement your own models, make sure that it is
+verified. Getting someone else to implement it is a good approach (following
+stdpopsim's QC procedure.)
+\item It's really important to do some basic sanity checking on your
+simulations. If your simulations are really big, do some small sims first and
+analysis on this to make that basic properties hold.
+\item Openness is essential. We only know about these errors because of
+open code and open source development processes. There must be many, many more
+out there.
+\item API design matters! A lot!!
+\end{itemize}
 

### 2020-05-15 paper.tex
martin bug

@@
 
-To illustrate the demography API, \msprime\ included a description of the
+To illustrate the demography API, \msprime\ included a description of a widely-used
 three population Out-of-Africa model~\citep{gutenkunst2009inferring}
@@
 
-[Description of what went wrong in the AJHG paper~\citep{martin2017human}.
-Some analysis of how this affects the conclusions of the paper.]
+In another publication using this model~\citep{martin2017human}, 
+another error crept in: the model itself was defined as suggested 
+in the documentation, and tested using model visualization and 
+debugging tools. After these checks were performed, however, the simulation 
+itself was finally performed with an incorrect parameter, leading to a vast 
+overestimate of the divergence across human populations.
 
-[How was the error found? I wonder if the reason that Alicia didn't
-do any basic checks on the simulation output was because it was
-so massive. Until recently we couldn't do things like compute Fst, etc
-on the tree sequences. Still, good enough reason to run small simulations
-first to make sure that you're doing something sensible.]
+In this case the error had a large effect on the simulation output and 
+the conclusions of the analysis. This simulation was performed to assess 
+the transferabilty of polygenic risk scores across human populations.
 
-This error was clearly caused by poor API design. By forcing the user to
-pass around three separate parameters to describe the demographic model,
-it's inviting simple mistakes such as this. In \msprime\ 1.0 we introduce
+Unfortunately, whereas the correct model suggests a mean $F_{ST}$ of 
+XXX across the three population, the simulated model generated an $F_{ST}$ of 
+XXX, exaggerating the difficulty in transferring genetic risk across populations (the difficulty still exists, Figure X).  The resulting
+publication has been influential in the discussion of health inequalities and 
+genomics, [[with over 350 citations since 2017 ?]].
+
+This particular error suggests three lessons.
+
+First, the msprime API design was poorly designed, requiring the user to pass
+three separate parameters to specify a demographic model. 
+ In \msprime\ 1.0 we introduce
 a Demography class, which wraps these three parameters.
@@
 ts = msprime.simulate(demography=demography)
-\end{lstlisting}
+\end{lstlisting}.
+
+Second, testing of the final simulated data is important. This can be challenging, 
+because the amount of simulated data can be large. However, recent progress in 
+computing summary statistics from tree sequence data \cite{} can make this easier. 
+Even if the entire simulation data cannot be easily evaluated, subsets of the data 
+can be examined to identify large error.
+
+Third, open data analysis pipelines are necessary for the self-correcting nature of science. 
+This subtle error of large effect was only discovered as one of us was trying to take
+advantage of the simulation pipeline developed in~\citep{martin2017human} to pursue 
+additional analyses on a similar topic. We identified the bug by tracing down an unexpected 
+result in our analysis.
 

### 2020-05-17 paper.tex
Edits throughout, include figures and analyses, comments, refs, etc

@@
 
+\usepackage{fullpage}
 \usepackage{listings}
+\usepackage{url}
+
+\usepackage{graphicx}
+\usepackage{color}
 
@@
 
+\newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
+
 \begin{document}
@@
 \abstract{
-We describe two errors made in defining population genetic models using the
-\msprime\ coalescent simulator.
-
+Simulation plays a central role in population genomics studies. Recent years
+have seen rapid improvements in software efficiency that make it possible to simulate 
+large genomic regions for many individuals sampled from large numbers of populations.
+The increase in complexity of possible demographic models also provides additional ways
+that we can get their implementation wrong. Here we describe two errors made in 
+defining population genetic models using the \msprime\ coalescent simulator that have 
+found their way into downstream analyses.
+We highlight a few ways that these errors have affected analyses based on these simulations
+and suggest measures that should be taken when simulating complicated demography.
 }
 
-\section*{Introduction}
+\section{Introduction}
 
@@
 of \msprime\ is the Python API that is its primary interface, greatly
-increasing the flexibility and efficiency over the standard approach of
+increasing the flexibility and ease of use over the standard approach of
 text-based command line interfaces. In particular, programs like \ms\
@@
 
-Describing these models of population history is still hard and error
-prone, however. In this note we discuss how two poor design decisions
+Implementing multi-population models of demographic history is still hard and error
+prone. In this note we discuss how two poor design decisions
 in \msprime's demography API lead to errors being made, which ended
 up in the scientific record. 
+\aprcomment{You could argue it is poor API design, and I see that.. but I also
+think there is something more to be said about best practices in implementing simulations, 
+including your own various verifications and sanity checks, and the need for robust
+implementation of commonly used models (stdpopsim). I do think that in the end
+the responsibility lies with the user in making sure that everything is implemented
+properly.}
 
@@
 three population Out-of-Africa model~\citep{gutenkunst2009inferring}
-as part of its tutorial documentation. This example model, however,
-was not correctly implemented.
+as part of its tutorial documentation. In this model (Fig.~\ref{fig:ooa_stats}A),
+Eurasian (CEU and CHB) and African (YRI) populations split from each other in the deep past, 
+followed by a more recent split of European and Asian populations, with variable rates of 
+continuous migration between each of the populations. However, the implementation in the 
+\msprime\ tutorial was incorrect. Namely, at the time of the split between African and Eurasian
+populations, migration between those two populations was allowed to continue for all time into the
+past, resulting in population structure when there should have been just a single 
+randomly mating ancestral population (Fig.~\ref{fig:ooa_stats}B).
 
-[What's wrong with the model, and how does this differ from the original. Some
-analysis.]
+Fortunately, the effect of this error is subtle. Population sizes and structure since the time of 
+the split are unaffected, so that differences in expected $F_{ST}$ are negligible between 
+the correct and incorrect model. However, the distribution of $T_{MRCA}$ in the deep past
+is distorted, and this results in an inflation of expected heterozygosity in each of the three
+populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).
 
@@
 
-Arguably, this error occured because of a poor API design choice.
-The MassMigration event is confusing and poorly explained. In
-\msprime\ 1.0 we're introducing a PopulationSplit event, which allows
-such models to be described more declaritively.
+Arguably, this error occurred because of a poor API design choice.
+Currently, to model a population split, the user must use a MassMigration event to move 
+lineage from one population to another and then also remember to turn off migration
+between those populations at the same time.
+The release of \msprime\ 1.0 will introduce a \texttt{PopulationSplit} event, which allows
+such models to be described more declaratively.
 
@@
 In another publication using this model~\citep{martin2017human}, 
-another error crept in: the model itself was defined as suggested 
-in the documentation, and tested using model visualization and 
-debugging tools. After these checks were performed, however, the simulation 
-itself was finally performed with an incorrect parameter, leading to a vast 
-overestimate of the divergence across human populations.
+a separate error was introduced: the model itself was defined as suggested 
+in the documentation (using somewhat more recently inferred parameters of that same
+model~\citep{gravel2011demographic}) and inspected using the \msprime\ 
+debugging tools. After these initial checks were made, however, the simulation 
+was performed without passing the parameter of demographic events,
+so that the three populations remain separated with low levels of migration and
+never merge as expected (Fig~\ref{fig:prs}A),
+leading to a vast overestimate of the divergence across human populations.
 
 In this case the error had a large effect on the simulation output and 
-the conclusions of the analysis. This simulation was performed to assess 
-the transferabilty of polygenic risk scores across human populations.
+the conclusions of the analysis. This simulation was performed to assess the 
+transferability of polygenic risk scores across human populations, in particular to
+explore how human demographic history and population structure affects our ability
+to predict genetic risk in populations of different ancestry than the originally studied
+population. The resulting publication has been influential in the discussion of 
+health inequalities and genomics, with over 350 citations since 2017.
 
-Unfortunately, whereas the correct model suggests a mean $F_{ST}$ of 
-XXX across the three population, the simulated model generated an $F_{ST}$ of 
-XXX, exaggerating the difficulty in transferring genetic risk across populations (the difficulty still exists, Figure X).  The resulting
-publication has been influential in the discussion of health inequalities and 
-genomics, [[with over 350 citations since 2017 ?]].
+Unfortunately, whereas the correct model predicts a mean $F_{ST}$ of 
+$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
+ranging between $0.3 - 0.6$, depending on the populations considered.
+This exaggerates the role that demography plays in limiting the transferability
+of genetic risk scores across populations. While this difficulty still exists 
+(Fig.~\ref{fig:prs}), risk prediction in each population is significantly improved 
+when using the correct demographic model (compare to Fig.~5 in~\citet{martin2017human}.
 
-This particular error suggests three lessons.
-
+These errors suggests three lessons. 
 First, the msprime API design was poorly designed, requiring the user to pass 
@@
 a Demography class, which wraps these three parameters.
-Thus, instread of writing
+Thus, instead of writing
 \begin{lstlisting}[frame=single]
@@
 ts = msprime.simulate(demography=demography)
-\end{lstlisting}.
+\end{lstlisting}
 
@@
 because the amount of simulated data can be large. However, recent progress in 
-computing summary statistics from tree sequence data \cite{} can make this easier. 
+computing summary statistics from tree sequence data can make this easier~\cite{ralph2020efficiently}. 
 Even if the entire simulation data cannot be easily evaluated, subsets of the data 
@@
 Third, open data analysis pipelines are necessary for the self-correcting nature of science. 
-This subtle error of large effect was only discovered as one of us was trying to take
-advantage of the simulation pipeline developed in~\citep{martin2017human} to pursue 
+This subtle error of large effect was only discovered through our own re-use of
+the simulation pipeline developed in~\citep{martin2017human} to pursue 
 additional analyses on a similar topic. We identified the bug by tracing down an unexpected 
-result in our analysis.
+result in our analysis. This error could not have been found and corrected without the open
+publication of the entire analysis pipeline from the original study.
 
@@
 
-Stuff to mention in no particular order:
+Defining demographic models is error prone, and incorrect implementation can  
+impact downstream analysis and interpretation. The errors that we highlighted here have 
+exposed some weaknesses of the current \msprime\ API, which will be fixed in an 
+upcoming release. Recognition of these difficulties is also a 
+primary motivating factor in efforts to standardize and vet demographic model 
+implementation in \stdpopsim~\cite{adrion2019community}, with quality-controlled 
+models and simulation resources for a growing number of commonly studied species.
+
+If you do need to implement your own demographic model, we recommend that steps are taken to 
+verify its correctness. A second pair of eyes (code review) or independent implementation 
+can go a long way in catching subtle bugs (this is the approach taken by \stdpopsim). It is also 
+very important to perform some basic statistical verification of the simulation, such as checking 
+output heterozygosity and $F_{ST}$ against expectations. If simulations are large so that 
+statistical validation is burdensome, you can instead simulate over a smaller genome length 
+and with fewer individuals to make sure basic properties hold. For example, a routine statistical 
+checks of $F_{ST}$ would surely have caught the error in~\citet{martin2017human}. 
+\aprcomment{too snarky...? but I think it's an important point}
+
+Finally, openness is essential. We only know about these errors because of open code and 
+open source development processes. \aprcomment{if we criticize the Martin paper for not
+checking basic statistics, we can at least give a kudos for putting the pipeline entirely on
+github here}. There must be many, many more mistakes out there, and an extra bit of care 
+is warranted to catch and correct such errors.
+
+\section{Methods}
+
 \begin{itemize}
-\item Defining demographic models is hard, and we need all the help we can get.
-   We really need \stdpopsim.
-\item If you do have to implement your own models, make sure that it is
-verified. Getting someone else to implement it is a good approach (following
-stdpopsim's QC procedure.)
-\item It's really important to do some basic sanity checking on your
-simulations. If your simulations are really big, do some small sims first and
-analysis on this to make that basic properties hold.
-\item Openness is essential. We only know about these errors because of
-open code and open source development processes. There must be many, many more
-out there.
-\item API design matters! A lot!!
+\item Figure 1: plotted using \texttt{demography} package (\url{https://github.com/apragsdale/demography}),
+allele frequency spectrum computed using \texttt{moments}~\cite{jouganous2017inferring}, and LD curves
+computed using \texttt{moments.LD}~\cite{ragsdale2019models}. $F_{ST}$ and other statistics computed
+from the expected AFS.
+\item Figure 2: original pipeline from \url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py},
+reran the experiment, with code available at \url{https://github.com/apragsdale/PRS}
 \end{itemize}
@@
 
+\pagebreak
+
+\begin{figure}[ht]
+\begin{center}
+\makebox[\textwidth][c]{\includegraphics{figures/ooa_expected_stats.pdf}}
+\caption{\textbf{Expected diversity statistics under the \citet{gutenkunst2009inferring} model}.
+    \textbf{(A)} The correctly implemented model.
+    \textbf{(B)} The incorrectly implemented model from the \msprime\ tutorial, with migration continuing
+    into the past.
+    \textbf{(C)} Marginal allele frequency spectra under the two models. Heterozygosity in the incorrect model
+    is inflated by $~3.5\%$, though the general shape of the distributions are qualitatively similar.
+    \textbf{(D)} Similarly, the increased heterozygosity leads to excess $D^2$, though the LD-decay is
+    qualitatively similar between models.
+    \textbf{(E-G)} True size history for each population plotted against the expected size history from 
+    the expected inverse coalescence rates.
+}
+\label{fig:ooa_stats}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht]
+\begin{center}
+\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
+\caption{\textbf{The transferability of PRS under neutrality}.
+    \textbf{(A)} In~\citet{martin2017human}, the simulated demographic model did not apply demographic
+    events in the past, so continental populations were simulated as isolated with low levels of migration
+    for all time.
+    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct 
+    demographic model and found that risk prediction was greatly improved over what was originally 
+    reported in all three populations.
+    \aprcomment{directly compare to figure 5 in the original paper}
+    \aprcomment{to do: fill in caption}
+}
+\label{fig:prs}
+\end{center}
+\end{figure}
+
 \end{document}

### 2020-05-17 analysis/models.py

@@ -1,0 +1,524 @@
+# Here we store the demographic models used in the analysis as functions that
+# return population configurations, demographic events, and migration matrices.
+# These are taken from the msprime documentation, before and after the mistake
+# in the model was found. They are copied exactly, except to set the choice
+# of printing the demography debugger as an argument.
+
+
+import msprime
+import math
+import demography
+import networkx as nx
+import moments, moments.LD as mold
+
+
+def out_of_africa_ancient_migration(print_debugger=False):
+    """
+    The out of africa model that does not turn off migration between demes 0 and 1
+    after the merger of the African and Eurasian populations.
+
+    print_debugger: If True, prints msprime demography debugger. Defaults to not print.
+
+    Returns population_configurations, migration_matrix, and demographic_events.
+    """
+    # First we set out the maximum likelihood values of the various parameters
+    # given in Table 1.
+    N_A = 7300
+    N_B = 2100
+    N_AF = 12300
+    N_EU0 = 1000
+    N_AS0 = 510
+    # Times are provided in years, so we convert into generations.
+    generation_time = 25
+    T_AF = 220e3 / generation_time
+    T_B = 140e3 / generation_time
+    T_EU_AS = 21.2e3 / generation_time
+    # We need to work out the starting (diploid) population sizes based on
+    # the growth rates provided for these two populations
+    r_EU = 0.004
+    r_AS = 0.0055
+    N_EU = N_EU0 / math.exp(-r_EU * T_EU_AS)
+    N_AS = N_AS0 / math.exp(-r_AS * T_EU_AS)
+    # Migration rates during the various epochs.
+    m_AF_B = 25e-5
+    m_AF_EU = 3e-5
+    m_AF_AS = 1.9e-5
+    m_EU_AS = 9.6e-5
+    # Population IDs correspond to their indexes in the population
+    # configuration array. Therefore, we have 0=YRI, 1=CEU and 2=CHB
+    # initially.
+    population_configurations = [
+        msprime.PopulationConfiguration(
+            sample_size=0, initial_size=N_AF),
[…]

### 2020-05-17 analysis/ooa_expected_statistics.py

@@ -1,0 +1,212 @@
+# Here we compute the AFS and LD decay curves for each population. To compute these
+# statistics, we use moments and use demography to define the OOA demographic models,
+# which also wraps the moments functions to compute the AFS or LD decay curves.
+# We also compute the coalescent rate trajectories for each model and compare to the
+# true population size history under the model. Finally, we print Fst for each pair
+# of populations under both the correct and incorrect model, as well as expected
+# heterozygosites in each population. Those values are summaries of the joint AFS.
+
+
+import msprime
+import models
+import demography
+import networkx as nx
+import math
+import moments, moments.LD as mold
+import numpy as np
+import matplotlib.pylab as plt
+import matplotlib
+import matplotlib.font_manager as fm
+from bokeh import palettes
+
+cb_colors = palettes.Colorblind8
+
+matplotlib.rcParams['text.usetex'] = True
+plt.rcParams['legend.title_fontsize'] = 'xx-small'
+matplotlib.rc('xtick', labelsize=6) 
+matplotlib.rc('ytick', labelsize=6) 
+matplotlib.rc('axes', labelsize=8) 
+matplotlib.rc('axes', titlesize=8) 
+matplotlib.rc('legend', fontsize=8)
+
+
+if __name__ == "__main__":
+    dg = models.ooa_dg()
+    dg_mig = models.ooa_mig_dg() # for plotting. we compute using a more stable function
+
+    ## get allele frequency spectra using moments (Jouganous et al 2017)
+    num_sam = 100 # number of samples per population
+    try:
+        fs = moments.Spectrum.from_file(f'data/ooa.{num_sam}.fs')
+    except IOError:
+        fs = dg.SFS(['YRI', 'CEU', 'CHB'], [num_sam, num_sam, num_sam])
+        fs.to_file(f'data/ooa.{num_sam}.fs')
+    try:
+        fs_mig = moments.Spectrum.from_file(f'data/ooa.ancient_mig.{num_sam}.fs')
+    except IOError:
[…]

### 2020-05-18 paper.tex
Some comments from Dominic on an early draft

@@
 \newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
+\newcommand{\dncomment}[1]{{\textcolor{red}{Dom: #1}}}
 
@@
 the large models we are using today, these are not comprehensible for humans.
+\dncomment{That's a pretty strong statement, might be good to give an example 
+or describe what it would look like (``specifying the gravel ooa model would take 
+10K lines of parameter specification''), or just say that its much less intuitive and 
+so more error prone}
 The Python API for \msprime\ is a great improvement, allowing the user to
@@
 Implementing multi-population models of demographic history is still hard and error
-prone. In this note we discuss how two poor design decisions
-in \msprime's demography API lead to errors being made, which ended
+prone. In this note we discuss how two poor
+\dncomment{Overly harsh I think, could drop this word} design decisions
+in \msprime's demography API lead
+\dncomment{Also not fair I don't think. I mean Alicia's bug hints at a lack of validation that 
+could have happened at any stage in the pipeline, and did happen elsewhere 
+I believe. Could make this more positive by describing how these design choices 
+could be improved to reduce the chance of errors going unnoticed/eliminate 
+some entirely} to errors being made, which ended
 up in the scientific record. 
@@
 many times it was copied? A quick search on github suggests around 30
-times].
+times]. \dncomment{I like this as a rough metric, like ``the number of copies 
+is unknown, but we found x copies on github''}
 
-Arguably, this error occurred because of a poor API design choice.
+Arguably, this error occurred because of a poor 
+\dncomment{Again I'd go positive. If msprime is so bad, god help the rest of us. 
+Something like ``To model a population split in the original API, ...''?}
+API design choice.
 Currently, to model a population split, the user must use a MassMigration event to move 
@@
 such models to be described more declaratively. 
+\dncomment{Strikes me as a bit programmer-centric. Maybe intuitively? Simply?}
 
@@
 These errors suggests three lessons. 
-First, the msprime API design was poorly designed, requiring the user to pass 
+First, the msprime API design was poorly designed, 
+\dncomment{Still don't like this word. Again if msprime is poor then most bioinformatics 
+software is a dumpster fire. What about ``...msprime API can be adapted to 
+simplify common use-cases''? A bit clunky but maybe along those lines
+}
+requiring the user to pass 
 three separate parameters to specify a demographic model. 
@@
 
-Defining demographic models is error prone, and incorrect implementation can  
+Defining demographic models is error prone \dncomment{Just a thought, but alternatively could say 
+``Demographic models are complex''}, and incorrect implementation can  
 impact downstream analysis and interpretation. The errors that we highlighted here have 
@@
 and with fewer individuals to make sure basic properties hold. For example, a routine statistical 
-checks of $F_{ST}$ would surely have caught the error in~\citet{martin2017human}. 
+checks of $F_{ST}$ would surely 
+\dncomment{I'd go with either this or 'routine' but maybe not both
+}
+have caught the error in~\citet{martin2017human}. 
 \aprcomment{too snarky...? but I think it's an important point}
@@
     for all time.
+    \dncomment{Might have been on your todos below but would help I think to have the correct model 
+    drawn out too (see now that you did in fig 1. I'd at least point to that to highlight the difference)}
     \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct 

### 2020-05-19 paper.tex
Minor tweaks.

@@
 \begin{lstlisting}[frame=single]
+demography = msprime.Demography(
+  population_configurations=population_configurations,
+  migration_matrix=migration_matrix,
+  demographic_events=demographic_events)
 dbg = demography.debug()
@@
 because the amount of simulated data can be large. However, recent progress in
-computing summary statistics from tree sequence data can make this easier~\cite{ralph2020efficiently}. 
+computing summary statistics from tree sequence data can make this
+easier~\citep{ralph2020efficiently}.
 Even if the entire simulation data cannot be easily evaluated, subsets of the data
@@
 primary motivating factor in efforts to standardize and vet demographic model
-implementation in \stdpopsim~\cite{adrion2019community}, with quality-controlled 
+implementation in \stdpopsim~\citep{adrion2019community}, with quality-controlled
 models and simulation resources for a growing number of commonly studied species.
@@
 \item Figure 1: plotted using \texttt{demography} package (\url{https://github.com/apragsdale/demography}),
-allele frequency spectrum computed using \texttt{moments}~\cite{jouganous2017inferring}, and LD curves
-computed using \texttt{moments.LD}~\cite{ragsdale2019models}. $F_{ST}$ and other statistics computed
+allele frequency spectrum computed using \texttt{moments}~\citep{jouganous2017inferring}, and LD curves
+computed using \texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other statistics computed
 from the expected AFS.

### 2020-05-19 paper.tex
edits throughout

--- 2020-05-17 paper.tex
@@
 \newcommand{\aprcomment}[1]{{\textcolor{blue}{APR: #1}}}
+\newcommand{\dncomment}[1]{{\textcolor{red}{Dom: #1}}}
+\newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
 
@@
 defining population genetic models using the \msprime\ coalescent simulator that have
-found their way into downstream analyses.
-We highlight a few ways that these errors have affected analyses based on these simulations
-and suggest measures that should be taken when simulating complicated demography.
+found their way in the published record.
+We discuss how these errors have affected analyses and suggest recommendations
+for software developers and users to reduce the risk of such errors.
 }
@@
 the large models we are using today, these are not comprehensible for humans.
+\dncomment{That's a pretty strong statement, might be good to give an example
+or describe what it would look like (``specifying the gravel ooa model would take
+10K lines of parameter specification''), or just say that its much less intuitive and
+so more error prone} \sgcomment{I think it's fine, esp. with JK's suggestions}
 The Python API for \msprime\ is a great improvement, allowing the user to
@@
 Implementing multi-population models of demographic history is still hard and error
-prone. In this note we discuss how two poor design decisions
-in \msprime's demography API lead to errors being made, which ended
-up in the scientific record. 
-\aprcomment{You could argue it is poor API design, and I see that.. but I also
-think there is something more to be said about best practices in implementing simulations, 
-including your own various verifications and sanity checks, and the need for robust
-implementation of commonly used models (stdpopsim). I do think that in the end
-the responsibility lies with the user in making sure that everything is implemented
-properly.}
+prone, however. In this note we discuss two implementation errors that found their way 
+in the scientific record.
+
+We identify two unfortunate design decisions in \msprime's demography API 
+that paved the way for these errors, and highlight
 
@@
 populations, migration between those two populations was allowed to continue for all time into the
-past, resulting in population structure when there should have been just a single 
-randomly mating ancestral population (Fig.~\ref{fig:ooa_stats}B).
+past. Even though the population split was implemented correcly, the specified demographic model had 
+an additional ghost population prior to the earliest split (Fig.~\ref{fig:ooa_stats}B).
 
@@
 the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
-the correct and incorrect model. However, the distribution of $T_{MRCA}$ in the deep past
-is distorted, and this results in an inflation of expected heterozygosity in each of the three
+the correct and incorrect model. However, the presence of a ghost population
+changes the distribution of $T_{MRCA}$ in the deep past
+is distorted and increases expected heterozygosity in each of the three
 populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).
@@
 This model was subsequently copied
-several times and used in papers [JK: do we want to estimate how
+several times and used in publications [JK: do we want to estimate how
 many times it was copied? A quick search on github suggests around 30
-times].
+times]. \dncomment{I like this as a rough metric, like ``the number of copies
+is unknown, but we found x copies on github''}
 
-Arguably, this error occurred because of a poor API design choice.
+Such errors could be prevented by improved API design choice.
 Currently, to model a population split, the user must use a MassMigration event to move
-lineage from one population to another and then also remember to turn off migration
+lineages from one population to another they must then also remember to turn off migration
 between those populations at the same time.
@@
 such models to be described more declaratively.
+\dncomment{Strikes me as a bit programmer-centric. Maybe intuitively? Simply?}
+\sgcomment{maybe something like "... \texttt{PopulationSplit}  event, which will link the changes in 
+the number of populations to changes in demographic parameters such as
+migration rates and effective population sizes"  }
 
@@
 a separate error was introduced: the model itself was defined as suggested
-in the documentation (using somewhat more recently inferred parameters of that same
-model~\citep{gravel2011demographic}) and inspected using the \msprime\ 
+in the documentation (using updated parameters~\citep{gravel2011demographic}) and inspected using the \msprime\
 debugging tools. After these initial checks were made, however, the simulation
@@
 leading to a vast overestimate of the divergence across human populations. 
+Whereas the correct model predicts a mean $F_{ST}$ of
+$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
+ranging between $0.3 - 0.6$, depending on the populations considered.
 
-In this case the error had a large effect on the simulation output and 
-the conclusions of the analysis. This simulation was performed to assess the 
+This simulation was performed to assess the
 transferability of polygenic risk scores across human populations, in particular to
@@
 to predict genetic risk in populations of different ancestry than the originally studied
-population. The resulting publication has been influential in the discussion of 
+population. 
+The excess divergence exaggerates the role that demography plays in limiting the transferability
+of genetic risk scores across populations. The resulting publication has been influential in the discussion of
 health inequalities and genomics, with over 350 citations since 2017.
-
-Unfortunately, whereas the correct model predicts a mean $F_{ST}$ of 
-$0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
-ranging between $0.3 - 0.6$, depending on the populations considered.
-This exaggerates the role that demography plays in limiting the transferability
-of genetic risk scores across populations. While this difficulty still exists 
+We note that while difficulties in transferability remain in the corrected model
 (Fig.~\ref{fig:prs}), risk prediction in each population is significantly improved
@@
 These errors suggests three lessons.
-First, the msprime API design was poorly designed, requiring the user to pass 
-three separate parameters to specify a demographic model. 
-In \msprime\ 1.0 we introduce
-a Demography class, which wraps these three parameters.
+First, the design of user interface and API for scientific software matters,
+and many bugs can be prevented by using more intuitive 
+interfaces. Whereas the original msprime required the user to pass
+three separate parameters to specify a demographic model, \msprime\ 1.0
+ introduces a Demography class, which wraps these three parameters.
 Thus, instead of writing
@@
   migration_matrix=migration_matrix,
-  demographic_events=demographic_events)
+  demographic_events=demographic_events),
 \end{lstlisting}
+with the error-inducing need to re-enter parameter information,  
 we would now write
 \begin{lstlisting}[frame=single]
+demography = msprime.Demography(
+  population_configurations=population_configurations,
+  migration_matrix=migration_matrix,
+  demographic_events=demographic_events)
 dbg = demography.debug()
@@
 because the amount of simulated data can be large. However, recent progress in
-computing summary statistics from tree sequence data can make this easier~\cite{ralph2020efficiently}. 
+computing summary statistics from tree sequence data can make this
+easier~\citep{ralph2020efficiently}.
 Even if the entire simulation data cannot be easily evaluated, subsets of the data
-can be examined to identify large error.
+can be examined to identify large errors.
 
@@
 
-Defining demographic models is error prone, and incorrect implementation can  
-impact downstream analysis and interpretation. The errors that we highlighted here have 
-exposed some weaknesses of the current \msprime\ API, which will be fixed in an 
-upcoming release. Recognition of these difficulties is also a 
-primary motivating factor in efforts to standardize and vet demographic model 
-implementation in \stdpopsim~\cite{adrion2019community}, with quality-controlled 
+The implementation of complex demographic models is error prone, and such errors can 
+have a large impact on downstream analysis and interpretation. 
+Recognition of these difficulties drove efforts to standardize and peer-review demographic model
+implementation in \stdpopsim~\citep{adrion2019community}, with quality-controlled
 models and simulation resources for a growing number of commonly studied species.
 
-If you do need to implement your own demographic model, we recommend that steps are taken to 
-verify its correctness. A second pair of eyes (code review) or independent implementation 
-can go a long way in catching subtle bugs (this is the approach taken by \stdpopsim). It is also 
-very important to perform some basic statistical verification of the simulation, such as checking 
-output heterozygosity and $F_{ST}$ against expectations. If simulations are large so that 
-statistical validation is burdensome, you can instead simulate over a smaller genome length 
-and with fewer individuals to make sure basic properties hold. For example, a routine statistical 
-checks of $F_{ST}$ would surely have caught the error in~\citet{martin2017human}. 
-\aprcomment{too snarky...? but I think it's an important point}
+The first bug discussed in this note was indeed identified through a peer-review effort of the 
+model implementation,  and it would have been unlikely to be identified by inspection of the 
+resulting data. The second bug would have been comparably easy to identify by inspection of the data, 
+but the missing parameter was easy to overlook.
 
-Finally, openness is essential. We only know about these errors because of open code and 
-open source development processes. \aprcomment{if we criticize the Martin paper for not
-checking basic statistics, we can at least give a kudos for putting the pipeline entirely on
-github here}. There must be many, many more mistakes out there, and an extra bit of care 
-is warranted to catch and correct such errors.
+We therefore recommend the following steps to ensure more robust simulations.
+First, we recommend that demographic models used in publications 
+are systematically verified by a second pair of eyes in code review. If the implementation is complex enough
+that code review is impractical, the risk that bugs are introduced is extremely high and the software 
+interface likely needs design improvements.
+
+Second, we recommend basic statistical verification of all simulated data. If simulations are so large that
+statistical validation is burdensome, subsets of the simulated data can be analyzed. 
+\sgcomment{Here again, software design can help by providing functions to compute 
+basic statistics of genetic diversity. [...]}. Such tests would have caught the error in~\citet{martin2017human}.
+
+Third, and most importantly, openness is essential. We only know about these errors because of open code and
+open source development processes. By making their entire pipeline available, \citet{martin2017human} not only 
+enabled other research teams to build upon their findings, but they also enlisted the entire community to review their code.
+
+There must be many, many more mistakes out there, and we need both pre- and post-publication vigilance from 
+users and developers to ensure the soundness of the large body of simulation-based analyses.
 
@@
 \begin{itemize}
-\item Figure 1: plotted using \texttt{demography} package (\url{https://github.com/apragsdale/demography}),
-allele frequency spectrum computed using \texttt{moments}~\cite{jouganous2017inferring}, and LD curves
-computed using \texttt{moments.LD}~\cite{ragsdale2019models}. $F_{ST}$ and other statistics computed
+\item Figure 1 used plotting routine from the \texttt{demography} package (\url{https://github.com/apragsdale/demography}),
+allele frequency spectrum computed using \texttt{moments}~\citep{jouganous2017inferring}, and LD curves
+computed using \texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other statistics computed
 from the expected AFS.
-\item Figure 2: original pipeline from \url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py},
-reran the experiment, with code available at \url{https://github.com/apragsdale/PRS}
+\item Figure 2 used the original pipeline from \url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py},
+and was rerun using corrected parameters, with code available at \url{https://github.com/apragsdale/PRS}
 \end{itemize}
@@
     the expected inverse coalescence rates.
+    \sgcomment{Not sure what the 1 in Figure 1b means. I would put the mass migration arrow in blue. }
 }
@@
     events in the past, so continental populations were simulated as isolated with low levels of migration
-    for all time.
+    for all time. The intended model is shown in Figure \ref{fig:ooa_stats}.
     \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
@@
     reported in all three populations.
-    \aprcomment{directly compare to figure 5 in the original paper}
+    \aprcomment{directly compare to figure 5 in the original paper} \sgcomment{If we }
     \aprcomment{to do: fill in caption}

### 2020-05-19 paper.tex
ghost clarification

@@
 populations, migration between those two populations was allowed to continue for all time into the
-past. Even though the population split was implemented correcly, the specified demographic model had 
-an additional ghost population prior to the earliest split (Fig.~\ref{fig:ooa_stats}B).
+past. Even though the population split was implemented correctly, the specified demographic model had 
+an additional population prior to the earliest split exchanging migrants with the ancestral population 
+(Fig.~\ref{fig:ooa_stats}B) but not contributing otherwise to present-day populations.
 
-Fortunately, the effect of this error is subtle. Population sizes and structure since the time of
+Fortunately, the effect of this additional 'ghost' population are subtle. Population sizes and structure since the time of
 the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
-the correct and incorrect model. However, the presence of a ghost population
-changes the distribution of $T_{MRCA}$ in the deep past
-is distorted and increases expected heterozygosity in each of the three
-populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).
+the correct and incorrect model. However, the presence of the ghost population increases
+the effective population size and thus affects the distribution of older $T_{MRCA}$ and the expected 
+heterozygosity in each of the three populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).
 

### 2020-05-19 paper.tex
ghost clarification

@@
 
-Fortunately, the effect of this additional 'ghost' population are subtle. Population sizes and structure since the time of
+Fortunately, the effect of this additional, 'ghost' population are subtle. Population sizes and structure since the time of
 the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
 the correct and incorrect model. However, the presence of the ghost population increases
-the effective population size and thus affects the distribution of older $T_{MRCA}$ and the expected 
+the effective population size and thus the distribution of $T_{MRCA}$ and the expected 
 heterozygosity in each of the three populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).

### 2020-05-19 paper.tex
added comment on MArtin et al

--- 2020-05-19 paper.tex
@@
 
+Finally, from a genetics perspective, the corrected simulations indicate that the accuracy of genetic risk scores
+is still substantially reduced in understudied populations, supporting the main conclusion of~\citep{martin2017human}. 
+However, the reduction is much less pronounced than reported. In particular, we did not observe large
+differences in mean risk prediction across populations, as presented in~\citep{martin2017human} \sgcomment{check}.
+Thus genetic drift alone does not appear to induce large biases in mean predicted risk across ethnicities, at least for 
+the polygenic architectures considered here. \sgcomment{This might need some analytical backup}
+
 \section{Conclusions}
@@
 \begin{center}
-\makebox[\textwidth][c]{\includegraphics{figures/ooa_expected_stats.pdf}}
-\caption{\textbf{Expected diversity statistics under the \citet{gutenkunst2009inferring} model}.
-    \textbf{(A)} The correctly implemented model.
-    \textbf{(B)} The incorrectly implemented model from the \msprime\ tutorial, with migration continuing
-    into the past.
-    \textbf{(C)} Marginal allele frequency spectra under the two models. Heterozygosity in the incorrect model
-    is inflated by $~3.5\%$, though the general shape of the distributions are qualitatively similar.
-    \textbf{(D)} Similarly, the increased heterozygosity leads to excess $D^2$, though the LD-decay is
-    qualitatively similar between models.
-    \textbf{(E-G)} True size history for each population plotted against the expected size history from
-    the expected inverse coalescence rates.
-    \sgcomment{Not sure what the 1 in Figure 1b means. I would put the mass migration arrow in blue. }
-}
-\label{fig:ooa_stats}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht]
-\begin{center}
 \makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}

### 2020-05-20 paper.tex
edits throughout, add analysis scripts and data

--- 2020-05-19 paper.tex
@@
 
+\usepackage{authblk}
 \usepackage{graphicx}
@@
 \newcommand{\stdpopsim}[0]{\texttt{stdpopsim}}
+\newcommand{\tskit}[0]{\texttt{tskit}}
 
@@
 \newcommand{\sgcomment}[1]{{\textcolor{red}{SG: #1}}}
+\newcommand{\jkcomment}[1]{{\textcolor{magenta}{JK: #1}}}
 
@@
 \title{Describing demographic models is hard}
-\author{A permutation of (AR, DN, JK, SG)}
+\author[ ]{A permutation of}
+\author[1]{Aaron P. Ragsdale}
+\author[1]{Dominic Nelson}
+\author[2]{Jerome Kelleher}
+\author[1]{Simon Gravel}
+\affil[1]{McGill University and Genome Qu\'{e}bec Innovation Centre,
+McGill University, Montr\'{e}al, Qu\'{e}bec, Canada}
+\affil[2]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
+University of Oxford, Oxford, United Kingdom}
 \maketitle
@@
 defining population genetic models using the \msprime\ coalescent simulator that have
-found their way in the published record.
+found their way into the published record.
 We discuss how these errors have affected analyses and suggest recommendations
@@
 required users to specify cryptic command line options to describe demographic
-models [JK: maybe include an example ms command line?]. Particularly for
-the large models we are using today, these are not comprehensible for humans.
-\dncomment{That's a pretty strong statement, might be good to give an example
-or describe what it would look like (``specifying the gravel ooa model would take
-10K lines of parameter specification''), or just say that its much less intuitive and
-so more error prone} \sgcomment{I think it's fine, esp. with JK's suggestions}
+models \jkcomment{maybe include an example ms command line?}. Particularly for
+the large models we use today, these are not intuitive or comprehensible for humans.
 The Python API for \msprime\ is a great improvement, allowing the user to
-state models in a documented and reproducible manner. [JK: maybe show
-the same model as described above in msprime notation?]
+state models in a documented and reproducible manner. \jkcomment{maybe show
+the same model as described above in msprime notation?}
 
 Implementing multi-population models of demographic history is still hard and error
-prone, however. In this note we discuss two implementation errors that found their way 
-in the scientific record.
-
-We identify two unfortunate design decisions in \msprime's demography API 
-that paved the way for these errors, and highlight
+prone, however. In this note we discuss two implementation errors that arose through
+unfortunate design decisions in \msprime's demography API and which then found their way
+into the scientific record. While one of these errors is not expected to have produced qualitative
+mistakes, the second lead to more severe ramifications \aprcomment{departures?}. In light of these
+implementation errors, we discuss improvements to \msprime's API motivated by these
+discoveries and best practices for implementing and simulating complex multi-population
+demography.
 
@@
 continuous migration between each of the populations. However, the implementation in the
-\msprime\ tutorial was incorrect. Namely, at the time of the split between African and Eurasian
-populations, migration between those two populations was allowed to continue for all time into the
-past. Even though the population split was implemented correctly, the specified demographic model had 
-an additional population prior to the earliest split exchanging migrants with the ancestral population 
-(Fig.~\ref{fig:ooa_stats}B) but not contributing otherwise to present-day populations.
+\msprime\ tutorial was incorrect. Namely, before the time of the split of African and Eurasian
+populations, when there should have been just a single randomly mating population, migration was
+allowed to occur between the ancestral population and a second population with size equal to
+the Eurasian bottleneck size for all time into the past (Fig.~\ref{fig:ooa_stats}B).
 
-Fortunately, the effect of this additional, 'ghost' population are subtle. Population sizes and structure since the time of
+%migration between those two populations was allowed to continue for all time into the
+%past. Even though the population split was implemented correctly, the specified demographic model had 
+%an additional population prior to the earliest split exchanging migrants with the ancestral population 
+%(Fig.~\ref{fig:ooa_stats}B) but not contributing otherwise to present-day populations.
+
+Fortunately, the effects of this error are subtle. Population sizes and structure since the time of
 the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
-the correct and incorrect model. However, the presence of the ghost population increases
-the effective population size and thus the distribution of $T_{MRCA}$ and the expected 
-heterozygosity in each of the three populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).
+the correct and incorrect model. However, the ancient structure distorts the distribution
+of $T_{MRCA}$ beyond the split date and increases
+heterozygosity in contemporary populations roughly $4\%$ over expectations
+from the correct model (Fig.~\ref{fig:ooa_stats}E-F).
+
+%However, the presence of the ghost population increases
+%the effective population size \aprcomment{not true - the effect is a bit different than just change in Ne}
+%and thus affects the distribution of older $T_{MRCA}$ and the expected 
+%heterozygosity in each of the three populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).
 
 This model was subsequently copied
-several times and used in publications [JK: do we want to estimate how
+several times and used in publications \jkcomment{do we want to estimate how
 many times it was copied? A quick search on github suggests around 30
-times]. \dncomment{I like this as a rough metric, like ``the number of copies
+times]} \dncomment{I like this as a rough metric, like ``the number of copies
 is unknown, but we found x copies on github''}
 
-Such errors could be prevented by improved API design choice.
-Currently, to model a population split, the user must use a MassMigration event to move
-lineages from one population to another they must then also remember to turn off migration
+Such errors can be prevented by improved API design choice.
+To model a population split currently in \msprime, a user must specify a
+\texttt{MassMigration} event that moves lineages from one population to another
+and then must also remember to turn off migration
 between those populations at the same time.
-The release of \msprime\ 1.0 will introduce a \texttt{PopulationSplit} event, which allows
-such models to be described more declaratively.
-\dncomment{Strikes me as a bit programmer-centric. Maybe intuitively? Simply?}
-\sgcomment{maybe something like "... \texttt{PopulationSplit}  event, which will link the changes in 
-the number of populations to changes in demographic parameters such as
-migration rates and effective population sizes"  }
+The release of \msprime~ver.~1.0 will introduce a \texttt{PopulationSplit} event,
+which more intuitively links the merger of lineages with appropriate changes in
+migration rates at the time of the split.
 
@@
 a separate error was introduced: the model itself was defined as suggested
-in the documentation (using updated parameters~\citep{gravel2011demographic}) and inspected using the \msprime\
-debugging tools. After these initial checks were made, however, the simulation
+in the documentation (using updated parameters
+from~\citet{gravel2011demographic}) and inspected
+using the \msprime\ debugging tools.
+After these initial checks were made, however, the simulation
 was performed without passing the parameter of demographic events,
-so that the three populations remain separated with low levels of migration and
-never merge as expected (Fig~\ref{fig:prs}A),
+so that the three populations remained separated with low levels of migration and
+never merged as expected (Fig~\ref{fig:prs}A),
 leading to a vast overestimate of the divergence across human populations. 
@@
 ranging between $0.3 - 0.6$, depending on the populations considered.
+Overall diversity was also strongly affected: expected heterozygosity was more than
+doubled in African populations but just $30-40\%$ the expected
+levels in Eurasian populations when compared to expectations from the correct model.
 
@@
 population.
-The excess divergence exaggerates the role that demography plays in limiting the transferability
-of genetic risk scores across populations. The resulting publication has been influential in the discussion of
+The excess divergence exaggerated the role that demography plays
+in limiting the transferability of genetic risk scores across populations.
+The resulting publication has been influential in the discussion of
 health inequalities and genomics, with over 350 citations since 2017.
@@
 (Fig.~\ref{fig:prs}), risk prediction in each population is significantly improved
-when using the correct demographic model (compare to Fig.~5 in~\citet{martin2017human}.
+when using the correct demographic model
+(compare to Fig.~5 in~\citet{martin2017human}).
 
@@
 First, the design of user interface and API for scientific software matters,
-and many bugs can be prevented by using more intuitive 
-interfaces. Whereas the original msprime required the user to pass
-three separate parameters to specify a demographic model, \msprime\ 1.0
- introduces a Demography class, which wraps these three parameters.
+and many bugs can be prevented by using more intuitive interfaces.
+Whereas the original \msprime\ required the user to pass
+three separate parameters to specify a demographic model, \msprime~ver.~1.0
+will introduce a Demography class, which wraps these three parameters.
 Thus, instead of writing
@@
   migration_matrix=migration_matrix,
-  demographic_events=demographic_events),
+  demographic_events=demographic_events)
 \end{lstlisting}
@@
 
-Second, testing of the final simulated data is important. This can be challenging,
-because the amount of simulated data can be large. However, recent progress in
-computing summary statistics from tree sequence data can make this
-easier~\citep{ralph2020efficiently}.
+Second, testing of the final simulated data is important.
+This can be challenging, because the amount of simulated data can be large.
+However, recent progress in computing summary statistics from tree
+sequence data can make this easier~\citep{ralph2020efficiently}.
 Even if the entire simulation data cannot be easily evaluated, subsets of the data
@@
 This subtle error of large effect was only discovered through our own re-use of
-the simulation pipeline developed in~\citep{martin2017human} to pursue
+the simulation pipeline developed in~\citet{martin2017human} to pursue
 additional analyses on a similar topic. We identified the bug by tracing down an unexpected
@@
 
+Finally, from a genetics perspective, the corrected simulations indicate that
+the accuracy of genetic risk scores is still substantially reduced in understudied
+populations (Fig.~\ref{fig:prs}E-G), supporting the main conclusion of~\citet{martin2017human}.
+However, the reduction is much less pronounced than reported.
+In particular, we do not observe large differences in mean risk prediction
+across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
+At least for the neutral polygenic architectures considered here, genetic drift alone
+does not appear to induce large biases in mean predicted risk across ethnicities.
+
 \section{Conclusions}
 
-The implementation of complex demographic models is error prone, and such errors can 
-have a large impact on downstream analysis and interpretation. 
-Recognition of these difficulties drove efforts to standardize and peer-review demographic model
+The implementation of complex demographic models is error prone, and such errors
+can have a large impact on downstream analysis and interpretation.
+Recognition of these difficulties drove efforts to peer-review demographic model
 implementation in \stdpopsim~\citep{adrion2019community}, with quality-controlled
@@
 The first bug discussed in this note was indeed identified through a peer-review effort of the 
-model implementation,  and it would have been unlikely to be identified by inspection of the 
-resulting data. The second bug would have been comparably easy to identify by inspection of the data, 
-but the missing parameter was easy to overlook.
+model implementation, and it is not easily identified by inspecting the resulting data.
+The second bug would have been comparably easy to identify by inspection
+of the data, but the missing parameter was easy to overlook.
 
@@
 First, we recommend that demographic models used in publications 
-are systematically verified by a second pair of eyes in code review. If the implementation is complex enough
-that code review is impractical, the risk that bugs are introduced is extremely high and the software 
-interface likely needs design improvements.
+are verified by a second pair of eyes in code review or an independent implementation.
+If the implementation is complex enough that code review is impractical, the risk
+that bugs are introduced is extremely high and additional caution and statistical
+verification is warranted.
+Second, we recommend basic statistical verification of all simulated data.
+If simulations are so large that statistical validation is burdensome,
+subsets of the simulated data can be analyzed~\citep{ralph2020efficiently}.
+Such tests would have caught the error in~\citet{martin2017human}.
 
-Second, we recommend basic statistical verification of all simulated data. If simulations are so large that
-statistical validation is burdensome, subsets of the simulated data can be analyzed. 
-\sgcomment{Here again, software design can help by providing functions to compute 
-basic statistics of genetic diversity. [...]}. Such tests would have caught the error in~\citet{martin2017human}.
-
-Third, and most importantly, openness is essential. We only know about these errors because of open code and
-open source development processes. By making their entire pipeline available, \citet{martin2017human} not only 
-enabled other research teams to build upon their findings, but they also enlisted the entire community to review their code.
-
-There must be many, many more mistakes out there, and we need both pre- and post-publication vigilance from 
-users and developers to ensure the soundness of the large body of simulation-based analyses.
+Third, openness is essential. We only know about these errors because of open code and
+open source development processes. By making their entire pipeline available,
+\citet{martin2017human} not only enabled other research teams to build upon their findings,
+but they made it possible for such errors to be found and corrected.
+There must be many, many more mistakes out there, and we need both
+pre- and post-publication vigilance from users and developers to ensure the
+soundness of the large body of simulation-based analyses.
 
@@
 
+We computed the expected allele frequency spectrum using
+\texttt{moments}~ver.~1.0.3~\citep{jouganous2017inferring}, and LD-decay curves using
+\texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other diversity statistics
+were computed from the expected AFS and verified using branch statistics from the
+tree sequence records of \msprime\ simulations~\citep{ralph2020efficiently}.
+Demographic models were plotted using the \texttt{demography} package
+(\url{https://github.com/apragsdale/demography}, ver.~0.0.3).
+We used the original pipeline from~\citet{martin2017human} available from
+\url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py}.
+We updated the pipeline to run with the correct demographic parameters and more
+recent versions of \msprime\ and \tskit\, and the updated pipeline is
+available at \url{https://github.com/apragsdale/PRS}.
+Data and python scripts to recreate Figures~1~and~2 can be found at
+\url{https://github.com/jeromekelleher/msprime-model-errors}.
+
+\section*{Acknowledgements}
+
 \begin{itemize}
-\item Figure 1 used plotting routine from the \texttt{demography} package (\url{https://github.com/apragsdale/demography}),
-allele frequency spectrum computed using \texttt{moments}~\citep{jouganous2017inferring}, and LD curves
-computed using \texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other statistics computed
-from the expected AFS.
-\item Figure 2 used the original pipeline from \url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py},
-and was rerun using corrected parameters, with code available at \url{https://github.com/apragsdale/PRS}
+\item Alicia Martin
+\item 
 \end{itemize}
@@
     \textbf{(B)} The incorrectly implemented model from the \msprime\ tutorial, with migration continuing
-    into the past.
+    into the past beyond the mass migration event with proportion $1$ from the ancestral population
+    to the bottleneck population.
     \textbf{(C)} Marginal allele frequency spectra under the two models. Heterozygosity in the incorrect model
@@
     the expected inverse coalescence rates.
-    \sgcomment{Not sure what the 1 in Figure 1b means. I would put the mass migration arrow in blue. }
 }
@@
     events in the past, so continental populations were simulated as isolated with low levels of migration
-    for all time. The intended model is shown in Figure \ref{fig:ooa_stats}.
+    for all time. The intended model is shown in Figure \ref{fig:ooa_stats}A.
     \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
-    demographic model and found that risk prediction was greatly improved over what was originally
-    reported in all three populations.
-    \aprcomment{directly compare to figure 5 in the original paper} \sgcomment{If we }
-    \aprcomment{to do: fill in caption}
+    demographic model.
+    While risk prediction in the African and East Asian population is reduced compared to the studied
+    European population, the reduction in prediction accuracy is not as hopeless as originally reported.
+    Notably, under a neutral polygenic architecture, we do not expect significant bias in inferred PRS
+    in any of the populations (C, D).
+    Correlations were computed over 100 simulation replicates.
+    For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.
 }

### 2020-05-20 analysis/get_fst_het_prs.py

@@ -1,0 +1,79 @@
+# This script runs some number of simulations under the correct
+# and incorrect demographic model parameterized in Gravel et al (2011).
+# We use branch statistics to compute heterozygosity in each population
+# and pairwise Fst between pops.
+
+# Usage: python get_fst_het_prs.py num_reps num_samples_per_population
+
+import models
+import msprime
+import tskit
+import sys
+
+
+if __name__ == "__main__":
+    reps, nsam = int(sys.argv[1]), int(sys.argv[2])
+
+    pc, mm, de = models.out_of_africa_gravel()
+    for i in range(3):
+        pc[i].sample_size = nsam
+
+    # run without demographic events
+    H0_inc = 0
+    H1_inc = 0
+    H2_inc = 0
+    F01_inc = 0
+    F02_inc = 0
+    F12_inc = 0
+
+    for _ in range(reps):
+        ts = msprime.simulate(
+            population_configurations=pc,
+            migration_matrix=mm,
+            recombination_rate=0.01,
+            length=1,
+        ) 
+        H0_inc += ts.diversity(range(nsam), mode='branch') 
+        H1_inc += ts.diversity(range(nsam, 2*nsam), mode='branch') 
+        H2_inc += ts.diversity(range(2*nsam, 3*nsam), mode='branch')
+        F01_inc += ts.Fst((range(nsam), range(nsam, 2*nsam)), mode='branch')
+        F02_inc += ts.Fst((range(nsam), range(2*nsam, 3*nsam)), mode='branch')
+        F12_inc += ts.Fst((range(nsam, 2*nsam), range(2*nsam, 3*nsam)), mode='branch')
+
+    # run with demographic events
+    H0_cor = 0
+    H1_cor = 0
+    H2_cor = 0
+    F01_cor = 0
+    F02_cor = 0
+    F12_cor = 0
+
+    for _ in range(reps):
+        ts = msprime.simulate(
+            population_configurations=pc,
+            migration_matrix=mm,
+            recombination_rate=0.01,
+            length=1,
+            demographic_events=de
+        ) 
+        H0_cor += ts.diversity(range(nsam), mode='branch') 
+        H1_cor += ts.diversity(range(nsam, 2*nsam), mode='branch') 
[…]

### 2020-05-20 analysis/ooa_expected_statistics.py

@@ -116,14 +116,14 @@ if __name__ == "__main__":
     demography.plotting.plot_demography(dg, leaf_order=['YRI', 'CEU', 'CHB'], ax=ax1,
                                         gen=25, stacked=[('A', 'YRI')], flipped=['CEU'],
                                         root_length=2, padding=3)
-    ax1.set_ylim(top=300)
+    ax1.set_ylim([0, 300])
     ax1.spines['left'].set_bounds(0, ax1.get_ylim()[1])
 
     ax2 = plt.subplot2grid((3, 6), (0, 3), colspan=3)
     demography.plotting.plot_demography(dg_mig, leaf_order=['YRI', 'CEU', 'CHB'], ax=ax2,
                                         gen=25, stacked=[('A', 'YRI')], flipped=['CEU'],
                                         padding=3)
-    ax2.set_ylim(top=300)
+    ax2.set_ylim([0, 300])
     ax2.spines['left'].set_bounds(0, ax2.get_ylim()[1])
     ax2.set_ylabel('')
 

### 2020-05-20 analysis/prs_plot.py

@@ -52,7 +52,7 @@ if __name__ == "__main__":
     pc, mm, de, dg = models.martin_model(merge_time=110000)
 
     # prs simulation data
-    fname = f'data/sim28_nhaps_400000_400000_400000_h2_0.67_m_1000_alpha_0.0.prs.gz'
+    fname = f'data/example_sim_nhaps_400000_400000_400000_h2_0.67_m_1000_alpha_0.0.prs.gz'
     data = pandas.read_csv(fname, sep='\t')
     AFR_true = data[data['Pop']=='AFR']['PRS_true']
     AFR_infer = data[data['Pop']=='AFR']['PRS_infer']

### 2020-05-20 paper.tex
Update readme with GitHub repos.

@@
 \title{Describing demographic models is hard}
-\author[ ]{A permutation of}
 \author[1]{Aaron P. Ragsdale}
 \author[1]{Dominic Nelson}
-\author[2]{Jerome Kelleher}
-\author[1]{Simon Gravel}
+\author[1,$\star$]{Simon Gravel}
+\author[2,$\star$]{Jerome Kelleher}
 \affil[1]{McGill University and Genome Qu\'{e}bec Innovation Centre,
@@
 University of Oxford, Oxford, United Kingdom}
+\affil[$\star$]{Joint senior authors, listed alphabetically}
 \maketitle
@@
 
-The \msprime\ coalescent simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
-is now quite widely used. The large increases in efficiency over the classical
+The \msprime\ coalescent
+simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
+is now widely used. The large increases in efficiency over the classical
 \ms\ program~\citep{hudson2002generating} make it feasible to simulate large
@@
 continuous migration between each of the populations. However, the implementation in the
-\msprime\ tutorial was incorrect. Namely, before the time of the split of African and Eurasian
+\msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
 populations, when there should have been just a single randomly mating population, migration was
@@
 
-%migration between those two populations was allowed to continue for all time into the
-%past. Even though the population split was implemented correctly, the specified demographic model had 
-%an additional population prior to the earliest split exchanging migrants with the ancestral population 
-%(Fig.~\ref{fig:ooa_stats}B) but not contributing otherwise to present-day populations.
+This model was subsequently copied
+several times and used in publications \jkcomment{do we want to estimate how
+many times it was copied? A quick search on github suggests around 30
+times]} \dncomment{I like this as a rough metric, like ``the number of copies
+is unknown, but we found x copies on github''}
+
+It was used in a number of publications
+~\citep{martin2017human,albers2020dating,kelleher2019inferring,tong2020population,hubisz2019mapping,
+overcast2019integrated,zhou2018popdemog,pfaffelhuber2020choose}
 
@@
 %heterozygosity in each of the three populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).
-
-This model was subsequently copied
-several times and used in publications \jkcomment{do we want to estimate how
-many times it was copied? A quick search on github suggests around 30
-times]} \dncomment{I like this as a rough metric, like ``the number of copies
-is unknown, but we found x copies on github''}
 

### 2020-05-20 paper.tex
More analysis of github repos.

@@
 This model was subsequently copied
-several times and used in publications \jkcomment{do we want to estimate how
+several times and used in publications. By searching for some keywords on
+GitHub, we found 33 repositories containing either direct copies of the
+erroneous model code, or code that was obviously derived from it.
+(We have opened issues on each of these repos to alert the authors.)
+
+\jkcomment{do we want to estimate how
 many times it was copied? A quick search on github suggests around 30

### 2020-05-20 paper.tex
Updated papers affected.

@@
 
-\jkcomment{do we want to estimate how
-many times it was copied? A quick search on github suggests around 30
-times]} \dncomment{I like this as a rough metric, like ``the number of copies
-is unknown, but we found x copies on github''}
-
 It was used in a number of publications
-~\citep{martin2017human,albers2020dating,kelleher2019inferring,tong2020population,hubisz2019mapping,
+~\citep{martin2017human,albers2020dating,kelleher2019inferring,tong2020population,
 overcast2019integrated,zhou2018popdemog,pfaffelhuber2020choose}

### 2020-05-20 paper.tex
Pass through the bad model section.

@@
 allowed to occur between the ancestral population and a second population with size equal to
-the Eurasian bottleneck size for all time into the past (Fig.~\ref{fig:ooa_stats}B).
+the Eurasian bottleneck size for all time into the past
+(Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
+the tutorial for \msprime\ version 0.3.0 and remained in the documentation for
+around four years.
 
-This model was subsequently copied
-several times and used in publications. By searching for some keywords on
-GitHub, we found 33 repositories containing either direct copies of the
-erroneous model code, or code that was obviously derived from it.
-(We have opened issues on each of these repos to alert the authors.)
-
-It was used in a number of publications
-~\citep{martin2017human,albers2020dating,kelleher2019inferring,tong2020population,
-overcast2019integrated,zhou2018popdemog,pfaffelhuber2020choose}
-
-Fortunately, the effects of this error are subtle. Population sizes and structure since the time of
+Fortunately, the effects of this error are subtle and it would take
+and exceedingly sensitive method to detect the difference.
+Population sizes and structure since the time of
 the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
@@
 
-%However, the presence of the ghost population increases
-%the effective population size \aprcomment{not true - the effect is a bit different than just change in Ne}
-%and thus affects the distribution of older $T_{MRCA}$ and the expected
-%heterozygosity in each of the three populations by up to $4\%$ (Fig.~\ref{fig:ooa_stats}C-G).
+It is fortunate that the error in the model description has quite a
+small effect because the tutorial code has been copied many times
+and used in publications. By searching for some identifying strings
+from the model definition on GitHub,
+we found 32 repositories containing either direct copies of the
+erroneous model code, or code that was obviously derived from it.
+(We have opened issues on each of these repos to alert the authors.)
+In most cases the publications use simulations to test some inference
+method which is not directly concerned with detailed demography,
+and the model is simply used as an example of a complex population
+history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
+\cite{zhou2018popdemog} use the incorrect model as an example
+of their method for visualising demographic models can support
+\msprime\ input.
+% TODO ask Franz about this
+Finally, \cite{pfaffelhuber2020choose}
+use simulations of the incorrect model demography to evalute
+their method for choosing ancestry informative markers.
+Given the very subtle effect of the incorrect
+model on demography (and the fact the method was evaluated using other
+simulations and real data), it seems unlikely that the model details
+will have any qualitative effect on their conclusions.
 
-Such errors can be prevented by improved API design choice.
+This long-standing error could have been prevented by better API design.
 To model a population split currently in \msprime, a user must specify a
@@
 between those populations at the same time.
-The release of \msprime~ver.~1.0 will introduce a \texttt{PopulationSplit} event,
+The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
 which more intuitively links the merger of lineages with appropriate changes in
@@
 \url{https://github.com/jeromekelleher/msprime-model-errors}.
+A full list of the GitHub repositories containing copies of the erroneous
+model are also given here.
 

### 2020-05-20 paper.tex
sharpen focus of Martin section and move material to conclusions

--- 2020-05-20 paper.tex
@@
 \maketitle
+
+\aprcomment{title suggestions:\\
+}
 
@@
 
-\section{A missing parameter}
+\section{A missing parameter (or) }
 
@@
 leading to a vast overestimate of the divergence across human populations.
-Whereas the correct model predicts a mean $F_{ST}$ of
+While the correct model predicts a mean $F_{ST}$ of
 $0.05 - 0.10$ across the three population, the simulated model generated $F_{ST}$
@@
 transferability of polygenic risk scores across human populations, in particular to
-explore how human demographic history and population structure affects our ability
+explore how human demographic history and population structure affect our ability
 to predict genetic risk in populations of different ancestry than the originally studied
 population.
-The excess divergence exaggerated the role that demography plays
-in limiting the transferability of genetic risk scores across populations.
 The resulting publication has been influential in the discussion of
 health inequalities and genomics, with over 350 citations since 2017.
-We note that while difficulties in transferability remain in the corrected model
-(Fig.~\ref{fig:prs}), risk prediction in each population is significantly improved
-when using the correct demographic model
-(compare to Fig.~5 in~\citet{martin2017human}).
+But while the first implementation error had a small effect on resulting diversity statistics,
+this error more strongly impacts simulated diversity and interpretations based on that data.
+The excess divergence under the incorrect model here exaggerated the role that demography
+plays in limiting the transferability of genetic risk scores across populations.
 
-These errors suggests three lessons.
+Difficulties in transferability remain in the corrected model (Fig.~\ref{fig:prs}), although
+risk prediction in each population is significantly improved when using the correct
+demographic model (compare to Fig.~5 in~\citet{martin2017human}).
+The corrected simulations indicate that the accuracy of genetic risk scores is still
+substantially reduced in understudied populations (Fig.~\ref{fig:prs}E-G),
+supporting the main conclusion of~\citet{martin2017human}.
+However, the reduction is much less pronounced than reported.
+In particular, we do not observe large differences in mean risk prediction
+across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
+At least for the neutral polygenic architectures considered here, genetic drift alone
+does not appear to induce large directional biases in mean predicted risk across ethnicities.
+
+\section{Conclusions}
+
+The implementation of complex demographic models is error prone, and such errors
+can have a large impact on downstream analysis and interpretation. 
+The discovery and correction of the simulation errors discussed here underscore
+how API design choices and 
+and we recommend the following steps to ensure more robust simulations.
+
 First, the design of user interface and API for scientific software matters,
@@
 
-Second, testing of the final simulated data is important.
-This can be challenging, because the amount of simulated data can be large.
-However, recent progress in computing summary statistics from tree
-sequence data can make this easier~\citep{ralph2020efficiently}.
-Even if the entire simulation data cannot be easily evaluated, subsets of the data
-can be examined to identify large errors.
-
-Third, open data analysis pipelines are necessary for the self-correcting nature of science.
-This subtle error of large effect was only discovered through our own re-use of
-the simulation pipeline developed in~\citet{martin2017human} to pursue
-additional analyses on a similar topic. We identified the bug by tracing down an unexpected
-result in our analysis. This error could not have been found and corrected without the open
-publication of the entire analysis pipeline from the original study.
-
-Finally, from a genetics perspective, the corrected simulations indicate that
-the accuracy of genetic risk scores is still substantially reduced in understudied
-populations (Fig.~\ref{fig:prs}E-G), supporting the main conclusion of~\citet{martin2017human}.
-However, the reduction is much less pronounced than reported.
-In particular, we do not observe large differences in mean risk prediction
-across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
-At least for the neutral polygenic architectures considered here, genetic drift alone
-does not appear to induce large biases in mean predicted risk across ethnicities.
-
-\section{Conclusions}
-
-The implementation of complex demographic models is error prone, and such errors
-can have a large impact on downstream analysis and interpretation.
 Recognition of these difficulties drove efforts to peer-review demographic model
@@
 verification is warranted.
+
 Second, we recommend basic statistical verification of all simulated data.
@@
 soundness of the large body of simulation-based analyses.
+
+----
+
+These errors suggests three lessons.
+
+Second, testing of the final simulated data is important. \aprcomment{verification of simulations, wrapped together}
+This can be challenging, because the amount of simulated data can be large.
+However, recent progress in computing summary statistics from tree
+sequence data can make this easier~\citep{ralph2020efficiently}.
+Even if the entire simulation data cannot be easily evaluated, subsets of the data
+can be examined to identify large errors.
+
+Third, open data analysis pipelines are necessary for the self-correcting nature of science.
+This subtle error of large effect was only discovered through our own re-use of
+the simulation pipeline developed in~\citet{martin2017human} to pursue
+additional analyses on a similar topic. We identified the bug by tracing down an unexpected
+result in our analysis. This error could not have been found and corrected without the open
+publication of the entire analysis pipeline from the original study.
 

### 2020-05-20 paper.tex

@@
 allowed to occur between the ancestral population and a second population with size equal to
-the Eurasian bottleneck size for all time into the past (Fig.~\ref{fig:ooa_stats}B).
+the Eurasian bottleneck size for all time into the past
+(Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
+the tutorial for \msprime\ version 0.3.0 and remained in the documentation for
+around four years.
 
-This model was subsequently copied
-several times and used in publications. By searching for some keywords on
-GitHub, we found 33 repositories containing either direct copies of the
-erroneous model code, or code that was obviously derived from it.
-(We have opened issues on each of these repos to alert the authors.)
-
-\jkcomment{do we want to estimate how
-many times it was copied? A quick search on github suggests around 30
-times]} \dncomment{I like this as a rough metric, like ``the number of copies
-is unknown, but we found x copies on github''}
-
-It was used in a number of publications
-~\citep{martin2017human,albers2020dating,kelleher2019inferring,tong2020population,hubisz2019mapping,
-overcast2019integrated,zhou2018popdemog,pfaffelhuber2020choose}
-
-Fortunately, the effects of this error are subtle. Population sizes and structure since the time of
+Fortunately, the effects of this error are subtle and it would take
+and exceedingly sensitive method to detect the difference.
+Population sizes and structure since the time of
 the split are unaffected, so that differences in expected $F_{ST}$ are negligible between

### 2020-05-20 paper.tex
clean up conclusions

@@
 Fortunately, the effects of this error are subtle and it would take
-and exceedingly sensitive method to detect the difference.
+an exceedingly sensitive method to detect the difference.
 Population sizes and structure since the time of
@@
 can have a large impact on downstream analysis and interpretation. 
-The discovery and correction of the simulation errors discussed here underscore
-how API design choices and 
-and we recommend the following steps to ensure more robust simulations.
+The discovery and correction of the demographic models discussed here underscore
+how API design choice can lead to the propagation of mistakes that are difficult
+to notice. We therefore recommend the following steps to ensure more robust simulations.
 
@@
 
-Recognition of these difficulties drove efforts to peer-review demographic model
-implementation in \stdpopsim~\citep{adrion2019community}, with quality-controlled
+Second, from a user's perspective, steps should be taken to validate the implemented
+demographic model. This could include verification through code review or an
+independent implementation of the model, which is the approach taken by
+\stdpopsim~\citep{adrion2019community} to compile quality-controlled
 models and simulation resources for a growing number of commonly studied species.
+Basic statistical validation should also be performed on any simulated model, and
+recent progress in computing summary statistics from tree sequence data can make
+this easier~\citep{ralph2020efficiently}. If simulations are so large that statistical
+validation is burdensome, subsets of the simulated data can be analyzed to ensure
+general diversity patterns are sound. Such tests likely would have caught the error
+in~\citet{martin2017human}.
 
-The first bug discussed in this note was indeed identified through a peer-review effort of the
-model implementation, and it is not easily identified by inspecting the resulting data.
-The second bug would have been comparably easy to identify by inspection
-of the data, but the missing parameter was easy to overlook.
-
-We therefore recommend the following steps to ensure more robust simulations.
-First, we recommend that demographic models used in publications
-are verified by a second pair of eyes in code review or an independent implementation.
-If the implementation is complex enough that code review is impractical, the risk
-that bugs are introduced is extremely high and additional caution and statistical
-verification is warranted.
-
-Second, we recommend basic statistical verification of all simulated data.
-If simulations are so large that statistical validation is burdensome,
-subsets of the simulated data can be analyzed~\citep{ralph2020efficiently}.
-Such tests would have caught the error in~\citet{martin2017human}.
-
-Third, openness is essential. We only know about these errors because of open code and
-open source development processes. By making their entire pipeline available,
+Finally, openness is essential to the self-correcting nature of science.
+We only know about these errors because of open code and
+open-source development processes. By making their entire pipeline available,
 \citet{martin2017human} not only enabled other research teams to build upon their findings,
@@
 soundness of the large body of simulation-based analyses.
-
-----
-
-These errors suggests three lessons.
-
-Second, testing of the final simulated data is important. \aprcomment{verification of simulations, wrapped together}
-This can be challenging, because the amount of simulated data can be large.
-However, recent progress in computing summary statistics from tree
-sequence data can make this easier~\citep{ralph2020efficiently}.
-Even if the entire simulation data cannot be easily evaluated, subsets of the data
-can be examined to identify large errors.
-
-Third, open data analysis pipelines are necessary for the self-correcting nature of science.
-This subtle error of large effect was only discovered through our own re-use of
-the simulation pipeline developed in~\citet{martin2017human} to pursue
-additional analyses on a similar topic. We identified the bug by tracing down an unexpected
-result in our analysis. This error could not have been found and corrected without the open
-publication of the entire analysis pipeline from the original study.
 

### 2020-05-20 paper.tex
header and title thoughts

@@
 
-\aprcomment{title suggestions:\\
+\aprcomment{title thoughts: \\
+Reducing errors in demographic model design and implementation \\
 }
@@
 
-\section{A missing parameter (or) }
+\section{Risk prediction in human populations}
 
@@
 doubled in African populations but just $30-40\%$ the expected
-levels in Eurasian populations when compared to expectations from the correct model.
+levels in Eurasian populations when compared to diversity under the correct model.
 
@@
 \item Alicia Martin
-\item
+\item stdpopsim as a whole or anyone in particular from that group?
+\item others?
 \end{itemize}

### 2020-05-20 paper.tex
edits throughout

@@
 
+\sgcomment{In the race to build more realistic and useful simulations of genetic diversity, 
+scientific software developers often focus on computational speed and biological realism. 
+User interface and application programming interfaces (API), however, also determine 
+whether software is used, and whether it is used correctly.
+ }
+
 The \msprime\ coalescent
 simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
-is now widely used. The large increases in efficiency over the classical
-\ms\ program~\citep{hudson2002generating} make it feasible to simulate large
+is now widely used. Much of its appeal is the large increase in efficiency over the classical
+\ms\ program~\citep{hudson2002generating} which make it feasible to simulate large
 samples of whole chromosomes for the first time. Another distinct advantage
@@
 state models in a documented and reproducible manner. \jkcomment{maybe show
-the same model as described above in msprime notation?}
+the same model as described above in msprime notation?}. \sgcomment{Other software, such as SLiM, 
+have developed advanced user interfaces and even dedicated programming languages 
+to facilitate the specification of various modes of evolution.\cite{}
+}
 
@@
 unfortunate design decisions in \msprime's demography API and which then found their way
-into the scientific record. While one of these errors is not expected to have produced qualitative
-mistakes, the second lead to more severe ramifications \aprcomment{departures?}. In light of these
-implementation errors, we discuss improvements to \msprime's API motivated by these
-discoveries and best practices for implementing and simulating complex multi-population
+into the scientific record. The first error has relatively mild effects on genetic diversity but was 
+used in many publications, while the second error was used only once but had a large impact on the 
+simulation results.  
+In light of these implementation errors, we discuss improvements to \msprime's API motivated by these
+discoveries and more general best practices for implementing and simulating complex multi-population
 demography.
@@
 
-Fortunately, the effects of this error are subtle and it would take
-an exceedingly sensitive method to detect the difference.
+Fortunately, the effects of this error are subtle.
 Population sizes and structure since the time of
@@
 the correct and incorrect model. However, the ancient structure distorts the distribution
-of $T_{MRCA}$ beyond the split date and increases
-heterozygosity in contemporary populations roughly $4\%$ over expectations
-from the correct model (Fig.~\ref{fig:ooa_stats}E-F).
+of early $T_{MRCAs}$. The extraneous ancient population increases increases the long-term effective 
+population size, leading to heterozygosity in contemporary populations
+increased by roughly $4\%$ over expectations from the correctly specified 
+model (Fig.~\ref{fig:ooa_stats}E-F).
 
@@
 erroneous model code, or code that was obviously derived from it.
-(We have opened issues on each of these repos to alert the authors.)
+(We have opened issues on each of these repositories to alert the authors.)
 In most cases the publications use simulations to test some inference
@@
 \cite{zhou2018popdemog} use the incorrect model as an example
-of their method for visualising demographic models can support
+of how their method for visualising demographic models can support
 \msprime\ input.
@@
 simulations and real data), it seems unlikely that the model details
-will have any qualitative effect on their conclusions.
+will have qualitative effect on their conclusions.
 
@@
 was performed without passing the parameter of demographic events,
-so that the three populations remained separated with low levels of migration and
-never merged as expected (Fig~\ref{fig:prs}A),
+so that the three populations
+never merged as expected and remained separated with low levels of migration   (Fig~\ref{fig:prs}A),
 leading to a vast overestimate of the divergence across human populations.
@@
 doubled in African populations but just $30-40\%$ the expected
-levels in Eurasian populations when compared to diversity under the correct model.
+levels in Eurasian populations when compared to diversity under the correctly specified model.
 
 This simulation was performed to assess the
-transferability of polygenic risk scores across human populations, in particular to
+transferability of polygenic risk scores across human populations. In other words, it sought to
 explore how human demographic history and population structure affect our ability
-to predict genetic risk in populations of different ancestry than the originally studied
-population.
-The resulting publication has been influential in the discussion of
+to predict genetic risk in diverse populations given the well-documented unequal representation
+in medical genetic studies \cite{Popejoy}. The resulting publication has been influential in the discussion of
 health inequalities and genomics, with over 350 citations since 2017.
-But while the first implementation error had a small effect on resulting diversity statistics,
-this error more strongly impacts simulated diversity and interpretations based on that data.
-The excess divergence under the incorrect model here exaggerated the role that demography
-plays in limiting the transferability of genetic risk scores across populations.
+
+The large excess in divergence under the incorrect model here exaggerated the role of demography
+and genetic drift in limiting the transferability of genetic risk scores across populations.
 
@@
 independent implementation of the model, which is the approach taken by
-\stdpopsim~\citep{adrion2019community} to compile quality-controlled
+\stdpopsim~\citep{adrion2019community} to build a library of quality-controlled
 models and simulation resources for a growing number of commonly studied species. 
+Such review likely would have caught the error in the msprime documentation.
+
 Basic statistical validation should also be performed on any simulated model, and

### 2020-05-20 paper.tex
title suggestions

--- 2020-05-20 paper.tex
@@
 Reducing errors in demographic model design and implementation \\
+Lessons learned from misspecified models of human history \\
+Lessons learned from buggy models of human history \\
+Lessons learned from bugs in population genetics models
 }

### 2020-05-20 paper.tex
title edits

--- 2020-05-20 paper.tex
@@
 
-\title{Describing demographic models is hard}
+\title{Lessons learned from bugs in models of human history}
 \author[1]{Aaron P. Ragsdale}
@@
 }
+\sgcomment{How to avoid simulating the wrong demographic model}
 
@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 
-\section{A bad tutorial example}
+\section*{Case 1: A misspecified model in msprime's documentation}
 
@@
 the correct and incorrect model. However, the ancient structure distorts the distribution
-of early $T_{MRCAs}$. The extraneous ancient population increases increases the long-term effective 
+of early $T_{MRCAs}$. The extraneous ancient population increases the long-term effective 
 population size, leading to heterozygosity in contemporary populations
@@
 
-\section{Risk prediction in human populations}
+\section*{Case 2: Incorrect model parameters in an analysis pipeline}
 
@@
 
-\section{Conclusions}
+\section*{Conclusions}
 
@@
 
-\section{Methods}
+\section*{Methods}
 

### 2020-05-20 paper.tex
Changes and refs

@@
 
-\title{Lessons learned from bugs in models of human history}
+\title{Lessons learned from buggy models of human history}
 \author[1]{Aaron P. Ragsdale}
@@
 Reducing errors in demographic model design and implementation \\
+Lessons learned from misspecified models of human history \\
+Lessons learned from buggy models of human history \\
+Lessons learned from bugs in population genetics models
 }
-\sgcomment{How to avoid simulating the wrong demographic model}
 
@@
 large genomic regions for many individuals sampled from large numbers of populations.
-The increase in complexity of possible demographic models also provides additional ways
+\aprcomment{fix this sentence ->} The increase in complexity of possible demographic models also provides additional ways
 that we can get their implementation wrong. Here we describe two errors made in
@@
 
-\section*{Case 1: A misspecified model in msprime's documentation}
+\section*{A bad tutorial example}
 
@@
 
-\section*{Case 2: Incorrect model parameters in an analysis pipeline}
+\section*{Risk prediction in human populations}
 
@@
 to predict genetic risk in diverse populations given the well-documented unequal representation
-in medical genetic studies \cite{Popejoy}. The resulting publication has been influential in the discussion of
-health inequalities and genomics, with over 350 citations since 2017.
-
+in medical genetic studies~\citep{popejoy2016genomics}. The resulting publication has been
+influential in the discussion of health inequalities and genomics, with over 350 citations since 2017.
 The large excess in divergence under the incorrect model here exaggerated the role of demography

### 2020-05-20 paper.tex

@@
 
-\title{Lessons learned from buggy models of human history}
+\title{Lessons learned from bugs in models of human history}
 \author[1]{Aaron P. Ragsdale}
@@
 
-\section*{A bad tutorial example}
+\section*{Case 1: A misspecified model in msprime's documentation}
 
@@
 
-\section*{Risk prediction in human populations}
+\section*{Case 2: Incorrect model parameters in an analysis pipeline}
 
@@
 
-\section*{Acknowledgements}
-
-\begin{itemize}
-\item Alicia Martin
-\item stdpopsim as a whole or anyone in particular from that group?
-\item others?
-\end{itemize}
-
 \bibliographystyle{plainnat}

### 2020-05-20 paper.tex
Edits to abstract and introduction, minor edits throughout

@@
 
-\aprcomment{title thoughts: \\
-Reducing errors in demographic model design and implementation \\
-Lessons learned from misspecified models of human history \\
-Lessons learned from buggy models of human history \\
-Lessons learned from bugs in population genetics models
-}
-\sgcomment{How to avoid simulating the wrong demographic model}
+%\aprcomment{title thoughts: \\
+%Reducing errors in demographic model design and implementation \\
+%Lessons learned from misspecified models of human history \\
+%Lessons learned from buggy models of human history \\
+%Lessons learned from bugs in population genetics models
+%}
+
+%\sgcomment{How to avoid simulating the wrong demographic model}
 
@@
 large genomic regions for many individuals sampled from large numbers of populations.
-\aprcomment{fix this sentence ->} The increase in complexity of possible demographic models also provides additional ways
-that we can get their implementation wrong. Here we describe two errors made in
-defining population genetic models using the \msprime\ coalescent simulator that have
-found their way into the published record.
-We discuss how these errors have affected analyses and suggest recommendations
+As the complexity of demographic models we study grows, there is an ever-increasing
+opportunity to introduce bugs in their implementation.
+Here we describe two errors made in defining population genetic models using
+the \msprime\ coalescent simulator that have found their way into the published record.
+We discuss how these errors have affected downstream analyses and give recommendations
 for software developers and users to reduce the risk of such errors.
@@
 
-\sgcomment{In the race to build more realistic and useful simulations of genetic diversity, 
+In the effort to build more realistic simulations of genetic diversity, 
 scientific software developers often focus on computational speed and biological realism. 
-User interface and application programming interfaces (API), however, also determine 
+Documentation and application programming interface (API), however, also determine 
 whether software is used, and whether it is used correctly.
- }
 
@@
 simulator~\citep{kelleher2016efficient,nelson2020accounting,kelleher2020coalescent}
-is now widely used. Much of its appeal is the large increase in efficiency over the classical
-\ms\ program~\citep{hudson2002generating} which make it feasible to simulate large
+is now widely used in genetics studies.
+Much of its appeal is the large increase in efficiency over the classical
+\ms\ program~\citep{hudson2002generating} which makes it feasible to simulate large
 samples of whole chromosomes for the first time. Another distinct advantage
@@
 increasing the flexibility and ease of use over the standard approach of
-text-based command line interfaces. In particular, programs like \ms\
-required users to specify cryptic command line options to describe demographic
-models \jkcomment{maybe include an example ms command line?}. Particularly for
-the large models we use today, these are not intuitive or comprehensible for humans.
+text-based command line interfaces.
+In particular, programs like \ms\ required users to specify cryptic command
+line options to describe demographic models.
+Particularly for the large models we use today, these are not
+intuitive or comprehensible for humans.
 The Python API for \msprime\ is a great improvement, allowing the user to
-state models in a documented and reproducible manner. \jkcomment{maybe show
-the same model as described above in msprime notation?}. \sgcomment{Other software, such as SLiM, 
-have developed advanced user interfaces and even dedicated programming languages 
-to facilitate the specification of various modes of evolution.\cite{}
-}
+state models in a documented and reproducible manner.
+\jkcomment{maybe show the OOA model as in both ms and msprime notation?}
+%\sgcomment{Other software, such as SLiM, 
+%have developed advanced user interfaces and even dedicated programming languages 
+%to facilitate the specification of various modes of evolution.\cite{}
+%}
 
 Implementing multi-population models of demographic history is still hard and error
-prone, however. In this note we discuss two implementation errors that arose through
+prone. In this note we discuss two implementation errors that arose through
 unfortunate design decisions in \msprime's demography API and which then found their way
@@
 (Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
-the tutorial for \msprime\ version 0.3.0 and remained in the documentation for
+the tutorial for \msprime\~ver.~0.3.0 and remained in the documentation for
 around four years.
@@
 Population sizes and structure since the time of
-the split are unaffected, so that differences in expected $F_{ST}$ are negligible between
+the split are unaffected, so differences in expected $F_{ST}$ are negligible between
 the correct and incorrect model. However, the ancient structure distorts the distribution
-of early $T_{MRCAs}$. The extraneous ancient population increases the long-term effective 
-population size, leading to heterozygosity in contemporary populations
-increased by roughly $4\%$ over expectations from the correctly specified 
-model (Fig.~\ref{fig:ooa_stats}E-F).
+of early TMRCAs. The extraneous ancient population increases the long-term effective 
+population size, resulting in roughly $4\%$ excess heterozygosity in 
+contemporary populations compared to the correctly specified model
+(Fig.~\ref{fig:ooa_stats}C-F).
 
@@
 (We have opened issues on each of these repositories to alert the authors.)
-In most cases the publications use simulations to test some inference
-method which is not directly concerned with detailed demography,
-and the model is simply used as an example of a complex population
+In most cases the publications used simulations to test some inference
+method and were not directly concerned with the details of the demography,
+and the model was simply used as an example of a complex population
 history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
-\cite{zhou2018popdemog} use the incorrect model as an example
+\cite{zhou2018popdemog} used the incorrect model as an example
 of how their method for visualising demographic models can support
@@
 Finally, \cite{pfaffelhuber2020choose}
-use simulations of the incorrect model demography to evalute
+used simulations of the incorrect model demography to evaluate
 their method for choosing ancestry informative markers.
@@
 simulations and real data), it seems unlikely that the model details
-will have qualitative effect on their conclusions.
+had any qualitative effect on their conclusions.
 
@@
 ranging between $0.3 - 0.6$, depending on the populations considered.
-Overall diversity was also strongly affected: expected heterozygosity was more than
+Overall diversity was also strongly affected: heterozygosity was more than
 doubled in African populations but just $30-40\%$ the expected
@@
 The implementation of complex demographic models is error prone, and such errors
-can have a large impact on downstream analysis and interpretation. 
+can have a large impact on downstream analyses and interpretation. 
 The discovery and correction of the demographic models discussed here underscore
@@
 and many bugs can be prevented by using more intuitive interfaces.
-Whereas the original \msprime\ required the user to pass
+Whereas the original \msprime\ API required the user to pass
 three separate parameters to specify a demographic model, \msprime~ver.~1.0
@@
 were computed from the expected AFS and verified using branch statistics from the
-tree sequence records of \msprime\ simulations~\citep{ralph2020efficiently}.
+tree sequence records of \msprime\ simulations using \tskit~ver.~0.2.3~\citep{ralph2020efficiently}.
 Demographic models were plotted using the \texttt{demography} package
@@
 We updated the pipeline to run with the correct demographic parameters and more
-recent versions of \msprime\ and \tskit\, and the updated pipeline is
+recent versions of \msprime\ and \tskit, and the updated pipeline is
 available at \url{https://github.com/apragsdale/PRS}.
@@
 \caption{\textbf{Expected diversity statistics under the \citet{gutenkunst2009inferring} model}.
-    \textbf{(A)} The correctly implemented model.
+    \textbf{(A)} The correctly implemented model. Dashed arrows depict continuous migration.
     \textbf{(B)} The incorrectly implemented model from the \msprime\ tutorial, with migration continuing

### 2020-05-20 paper.tex
Add ms and msprime OOA models

@@
 state models in a documented and reproducible manner.
-\jkcomment{maybe show the OOA model as in both ms and msprime notation?}
 %\sgcomment{Other software, such as SLiM, 
@@
 
+For reference, the~\citet{gutenkunst2009inferring} demographic model, as written in \ms, is
+\begin{lstlisting}[frame=single]
+-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
+-eg 0 2 116.010723 -eg 0 3 160.246047
+-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
+-ej 0.028985 3 2 -en 0.028985 2 0.287184
+-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
+-ej 0.197963 2 1 -en 0.303501 1 1
+\end{lstlisting}
+The equivalent model using \msprime\ is more human readable and provided in full at
+\url{https://github.com/jeromekelleher/msprime-model-errors}. The demographic events
+are given as
+\begin{lstlisting}[frame=single]
+demographic_events = [
+    # CEU and CHB merge into B with rate changes at T_EU_AS
+    msprime.MassMigration(
+        time=T_EU_AS, source=2, destination=1, proportion=1.0),
+    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
+    msprime.MigrationRateChange(
+        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
+    msprime.MigrationRateChange(
+        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
+    msprime.PopulationParametersChange(
+        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
+    # Population B merges into YRI at T_B
+    msprime.MassMigration(
+        time=T_B, source=1, destination=0, proportion=1.0),
+    msprime.MigrationRateChange(time=T_B, rate=0),
+    # Size changes to N_A at T_AF
+    msprime.PopulationParametersChange(
+        time=T_AF, initial_size=N_A, population_id=0)
+]
+\end{lstlisting}
+
 \bibliographystyle{plainnat}

### 2020-05-20 paper.tex
small edits and added comment on graphical inspectin

@@
 
-%\sgcomment{How to avoid simulating the wrong demographic model}
+\sgcomment{How to avoid simulating the wrong demographic model}
 
@@
 large genomic regions for many individuals sampled from large numbers of populations.
-As the complexity of demographic models we study grows, there is an ever-increasing
+As the complexity of demographic models we study grows, however, there is an ever-increasing
 opportunity to introduce bugs in their implementation.
@@
 intuitive or comprehensible for humans.
-The Python API for \msprime\ is a great improvement, allowing the user to
-state models in a documented and reproducible manner.
+The Python API for \msprime\, by contrast, allows the user to
+state models in a more human-readable manner: 
+Example of model specification for the two softwares is provided in the Methods section. 
 %\sgcomment{Other software, such as SLiM, 
@@
 
-Implementing multi-population models of demographic history is still hard and error
-prone. In this note we discuss two implementation errors that arose through
+Implementing multi-population models of demographic history for either software is hard and error-prone. 
+In this note we discuss two implementation errors that arose through
 unfortunate design decisions in \msprime's demography API and which then found their way
@@
 In light of these implementation errors, we discuss improvements to \msprime's API motivated by these
-discoveries and more general best practices for implementing and simulating complex multi-population
+discoveries and, more generally, best practices for implementing and simulating complex multi-population
 demography.
@@
 (Fig.~\ref{fig:ooa_stats}B). This incorrect model was introduced into
-the tutorial for \msprime\~ver.~0.3.0 and remained in the documentation for
+the tutorial for \msprime~ver.~0.3.0 and remained in the documentation for
 around four years.
@@
 Population sizes and structure since the time of
-the split are unaffected, so differences in expected $F_{ST}$ are negligible between
-the correct and incorrect model. However, the ancient structure distorts the distribution
+the earliest split are unaffected, so differences in expected $F_{ST}$ are negligible between
+the intended and incorrect models. However, the ancient structure distorts the distribution
 of early TMRCAs. The extraneous ancient population increases the long-term effective 
 population size, resulting in roughly $4\%$ excess heterozygosity in 
-contemporary populations compared to the correctly specified model
+contemporary populations compared to the intended model
 (Fig.~\ref{fig:ooa_stats}C-F).
 
-It is fortunate that the error in the model description has quite a
-small effect because the tutorial code has been copied many times
-and used in publications. By searching for some identifying strings
+Even though the error had a limited effect on simulated data, the tutorial code has been 
+copied many times and used in publications. By searching for some identifying strings
 from the model definition on GitHub,
@@
 Overall diversity was also strongly affected: heterozygosity was more than
-doubled in African populations but just $30-40\%$ the expected
-levels in Eurasian populations when compared to diversity under the correctly specified model.
+doubled in African populations and reduced by more than half in Eurasian populations 
+relative to the intended model.
 
@@
 
-Difficulties in transferability remain in the corrected model (Fig.~\ref{fig:prs}), although
-risk prediction in each population is significantly improved when using the correct
-demographic model (compare to Fig.~5 in~\citet{martin2017human}).
-The corrected simulations indicate that the accuracy of genetic risk scores is still
+Difficulties in transferability remain in the intended model (Fig.~\ref{fig:prs}), although
+risk prediction in each population is significantly improved (compare to Fig.~5 in~\citet{martin2017human}).
+The intended simulations indicate that the accuracy of genetic risk scores is still
 substantially reduced in understudied populations (Fig.~\ref{fig:prs}E-G),
@@
 in~\citet{martin2017human}.
+
+Graphical inspection of demographic models can also help identify issues, especially if 
+visualization can be automated. msprime does contain a demography debugging tool summarizing 
+demographic events occurring over time, and various efforts are underway to facilitate 
+the specification and visualization of population genetics models \cite{zhou2018popdemog}, \sgcomment{Mention Aaron's code and Slim?}.
 

### 2020-05-20 paper.tex
Minor language tweaks

@@
 state models in a more human-readable manner: 
-Example of model specification for the two softwares is provided in the Methods section. 
+an examplle of model specification for the two softwares is provided in the Methods section. 
 %\sgcomment{Other software, such as SLiM, 
@@
 followed by a more recent split of European and Asian populations, with variable rates of
-continuous migration between each of the populations. However, the implementation in the
+continuous migration between each of the populations. Regrettably, the implementation in the
 \msprime\ tutorial was incorrect. Before the time of the split of African and Eurasian
@@
 using the \msprime\ debugging tools.
-After these initial checks were made, however, the simulation
+Despite these initial checks being made, the simulation
 was performed without passing the parameter of demographic events,
@@
 will introduce a Demography class, which wraps these three parameters.
-Thus, instead of writing
+Instead of writing
 \begin{lstlisting}[frame=single]
@@
 Graphical inspection of demographic models can also help identify issues, especially if 
-visualization can be automated. msprime does contain a demography debugging tool summarizing 
+visualization can be automated. \msprime\ does contain a demography debugging tool summarizing 
 demographic events occurring over time, and various efforts are underway to facilitate 
-the specification and visualization of population genetics models \cite{zhou2018popdemog}, \sgcomment{Mention Aaron's code and Slim?}.
+the specification and visualization of population genetics models, such as~\citet{zhou2018popdemog}
+and the \texttt{demography} package used in this paper.
 

### 2020-05-21 paper.tex
Remove a few comments
affiliations

@@
 \author[2,$\star$]{Jerome Kelleher}
-\affil[1]{McGill University and Genome Qu\'{e}bec Innovation Centre,
+\affil[1]{McGill Genome Centre and Department of Human Genetics,
 McGill University, Montr\'{e}al, Qu\'{e}bec, Canada}
@@
 \maketitle
-
-%\aprcomment{title thoughts: \\
-%Reducing errors in demographic model design and implementation \\
-%Lessons learned from misspecified models of human history \\
-%Lessons learned from buggy models of human history \\
-%Lessons learned from bugs in population genetics models
-%}
-
-\sgcomment{How to avoid simulating the wrong demographic model}
 

### 2020-05-21 paper.tex
Round of edits.

@@
 \author[2,$\star$]{Jerome Kelleher}
-\affil[1]{McGill Genome Centre and Department of Human Genetics,
+\affil[1]{McGill University and Genome Qu\'{e}bec Innovation Centre,
 McGill University, Montr\'{e}al, Qu\'{e}bec, Canada}
@@
 large genomic regions for many individuals sampled from large numbers of populations.
-As the complexity of demographic models we study grows, however, there is an ever-increasing
+As the complexity of the demographic models we study grows, however, there is an ever-increasing
 opportunity to introduce bugs in their implementation.
@@
 samples of whole chromosomes for the first time. Another distinct advantage
-of \msprime\ is the Python API that is its primary interface, greatly
+of \msprime\ is its Python API, greatly
 increasing the flexibility and ease of use over the standard approach of
 text-based command line interfaces.
-In particular, programs like \ms\ required users to specify cryptic command
+In particular, programs like \ms\ require users to specify cryptic command
 line options to describe demographic models.
@@
 intuitive or comprehensible for humans.
-The Python API for \msprime\, by contrast, allows the user to
-state models in a more human-readable manner: 
-an examplle of model specification for the two softwares is provided in the Methods section. 
+The Python API for \msprime, by contrast, allows the user to
+state models in a more human-readable manner.
+An example contrasting the different approaches to model specification
+is provided in the Methods section.
 %\sgcomment{Other software, such as SLiM,
@@
 
-Implementing multi-population models of demographic history for either software is hard and error-prone. 
+Implementing multi-population models of demographic history is difficult and error-prone.
 In this note we discuss two implementation errors that arose through
@@
 
-Even though the error had a limited effect on simulated data, the tutorial code has been 
+Even though the error has a limited effect on simulated data, the tutorial code has been
 copied many times and used in publications. By searching for some identifying strings
@@
 (We have opened issues on each of these repositories to alert the authors.)
-In most cases the publications used simulations to test some inference
-method and were not directly concerned with the details of the demography,
-and the model was simply used as an example of a complex population
+In most cases the publications used simulations to test a non-demographic inference
+method, and the model was used as an example of a complex population
 history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
@@
 The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
-which more intuitively links the merger of lineages with appropriate changes in
+which more intuitively links the movement of lineages with appropriate changes in
 migration rates at the time of the split.
@@
 Despite these initial checks being made, the simulation
-was performed without passing the parameter of demographic events,
+was performed without passing the parameter defining demographic events,
 so that the three populations
-never merged as expected and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
+never merged and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
 leading to a vast overestimate of the divergence across human populations.
@@
 
+This error was easy to make, and could have been prevented by better API design.
+The original \msprime\ API requires the user to pass
+three separate parameters to specify a demographic model. For example,
+\begin{lstlisting}[frame=single]
+dbg = msprime.DemographyDebugger(
+  population_configurations=population_configurations,
+  demographic_events=demographic_events,
+  migration_matrix=migration_matrix)
+dbg.print_history()
+ts = msprime.simulate(
+  population_configurations=population_configurations,
+  demographic_events=demographic_events, # Missing in Martin et al. 2017!
+  migration_matrix=migration_matrix)
+\end{lstlisting}
+Note that the same three parameters must be passed to both the debugger
+and the simulate function.
+To help prevent such errors,
+\msprime~ver.~1.0 will introduce a Demography class. The above
+snippet would then be rewritten as
+\begin{lstlisting}[frame=single]
+demography = msprime.Demography(
+  population_configurations=population_configurations,
+  demographic_events=demographic_events,
+  migration_matrix=migration_matrix)
+dbg = demography.debug()
+dbg.print_history()
+ts = msprime.simulate(demography=demography)
+\end{lstlisting}
+This simple change to the interface makes it much less likely
+that different models are passed to the debugger and simulator,
+reducing the potential for error.
+
 \section*{Conclusions}
@@
 
-First, the design of user interface and API for scientific software matters,
-and many bugs can be prevented by using more intuitive interfaces.
-Whereas the original \msprime\ API required the user to pass
-three separate parameters to specify a demographic model, \msprime~ver.~1.0
-will introduce a Demography class, which wraps these three parameters.
-Instead of writing
-\begin{lstlisting}[frame=single]
-dbg = msprime.DemographyDebugger(
-  population_configurations=population_configurations,
-  migration_matrix=migration_matrix,
-  demographic_events=demographic_events)
-dbg.print_history()
-ts = msprime.simulate(
-  population_configurations=population_configurations,
-  migration_matrix=migration_matrix,
-  demographic_events=demographic_events)
-\end{lstlisting}
-with the error-inducing need to re-enter parameter information,
-we would now write
-\begin{lstlisting}[frame=single]
-demography = msprime.Demography(
-  population_configurations=population_configurations,
-  migration_matrix=migration_matrix,
-  demographic_events=demographic_events)
-dbg = demography.debug()
-dbg.print_history()
-ts = msprime.simulate(demography=demography)
-\end{lstlisting}
+Firstly, if possible, we recommend avoiding the implementation of
+demographic models entirely by using \stdpopsim~\citep{adrion2019community}.
+This is a ``standard library" of quality-controlled
+models and simulation resources for a growing number of commonly studied species.
+The resource is built around an open-source community development model
+with rigorous code review and quality control procedures.
+For example, when a contributor adds a new model it is not fully integrated
+into the catalog until a second, entirely independent, implementation of the
+model is provided by another developer. If the original and ``QC" versions
+of the model are precisely equal, then we can have a reasonable degree
+of confidence in correctness. It is through this QC process that we discovered
+the misspecified model in \msprime's documentation.
+If the required model is not present in \stdpopsim, then particular care should
+be taken to validate the implementation.
+This could include verification through code review or comparison with an
+independent implementation of the model. If the model is of wider interest to
+the community, we would encourage developers to contribute it to \stdpopsim:
+at the very least, the quality-control procedures in place provide a strong
+reassurance of correctness.
 
-Second, from a user's perspective, steps should be taken to validate the implemented
-demographic model. This could include verification through code review or an
-independent implementation of the model, which is the approach taken by
-\stdpopsim~\citep{adrion2019community} to build a library of quality-controlled
-models and simulation resources for a growing number of commonly studied species. 
-Such review likely would have caught the error in the msprime documentation.
-
-Basic statistical validation should also be performed on any simulated model, and
-recent progress in computing summary statistics from tree sequence data can make
-this easier~\citep{ralph2020efficiently}. If simulations are so large that statistical
-validation is burdensome, subsets of the simulated data can be analyzed to ensure
-general diversity patterns are sound. Such tests likely would have caught the error
-in~\citet{martin2017human}.
-
+Secondly, regardless of whether the model has been implemented
+locally or obtained from a resource like \stdpopsim,
+basic statistical validation should \emph{always} be performed on the
+results. Errors are all too easy to make, and an analysis of the
+basic statistical properties of the simulations is essential
+due diligence. For example, a brief analysis of $F_{ST}$ values
+would likely would have prevented the error in~\citet{martin2017human}.
+It is understandable that such analyses were not undertaken in this
+case, as processing such a huge dataset (600,000 samples) was
+highly challenging. Recent progress has made statistical analysis this scale much
+easier~\citep{ralph2020efficiently}, but the problems would have been
+apparent if smaller pilot simulations were undertaken and analysed.
 Graphical inspection of demographic models can also help identify issues, especially if
-visualization can be automated. \msprime\ does contain a demography debugging tool summarizing 
-demographic events occurring over time, and various efforts are underway to facilitate 
+visualization can be automated. The demography debugging tool in \msprime\
+summarizes demographic events occurring over time, and various efforts are underway to facilitate
 the specification and visualization of population genetics models, such as~\citet{zhou2018popdemog}
@@
 \citet{martin2017human} not only enabled other research teams to build upon their findings,
-but they made it possible for such errors to be found and corrected.
+but they made it possible for errors to be found and corrected.
 There must be many, many more mistakes out there, and we need both
@@
 \texttt{moments.LD}~\citep{ragsdale2019models}. $F_{ST}$ and other diversity statistics
-were computed from the expected AFS and verified using branch statistics from the
-tree sequence records of \msprime\ simulations using \tskit~ver.~0.2.3~\citep{ralph2020efficiently}.
+were computed from the expected AFS and verified using branch statistics from
+the output of \msprime\ simulations using \tskit~ver.~0.2.3~\citep{ralph2020efficiently}.
 Demographic models were plotted using the \texttt{demography} package
@@
 available at \url{https://github.com/apragsdale/PRS}.
-Data and python scripts to recreate Figures~1~and~2 can be found at
+Data and Python scripts to recreate Figures~1~and~2 can be found at
 \url{https://github.com/jeromekelleher/msprime-model-errors}.
@@
 \end{lstlisting}
+It is worth noting that the~\citet{gutenkunst2009inferring} model is relatively
+simple, and models with many more populations and parameters are increasingly
+common.
 

### 2020-05-21 paper.tex
discussion of Martin et al.

@@
 substantially reduced in understudied populations (Fig.~\ref{fig:prs}E-G),
-supporting the main conclusion of~\citet{martin2017human}.
+supporting one of the main conclusions of~\citet{martin2017human}.
 However, the reduction is much less pronounced than reported.
-In particular, we do not observe large differences in mean risk prediction
+In particular, we do not observe large differences in mean predicted risk 
 across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
 At least for the neutral polygenic architectures considered here, genetic drift alone
-does not appear to induce large directional biases in mean predicted risk across ethnicities.
+does not appear to induce large directional biases in mean predicted risk.
 

### 2020-05-22 paper.tex
Some edits after comments

@@
 the earliest split are unaffected, so differences in expected $F_{ST}$ are negligible between
-the intended and incorrect models. However, the ancient structure distorts the distribution
-of early TMRCAs. The extraneous ancient population increases the long-term effective
+the correct and incorrect models. However, the ancient structure distorts the distribution
+of early TMRCAs (Fig.~\ref{fig:ooa_stats}E-G).
+The extraneous ancient population increases the long-term effective
 population size, resulting in roughly $4\%$ excess heterozygosity in
-contemporary populations compared to the intended model
-(Fig.~\ref{fig:ooa_stats}C-F).
+contemporary populations compared to the intended model, but the overall effects
+on patterns of diversity are minimal (Fig.~\ref{fig:ooa_stats}C,D).
 
@@
 doubled in African populations and reduced by more than half in Eurasian populations
-relative to the intended model.
+relative to the correct model.
 
@@
 
-Difficulties in transferability remain in the intended model (Fig.~\ref{fig:prs}), although
+Difficulties in transferability remain in the corrected model (Fig.~\ref{fig:prs}), although
 risk prediction in each population is significantly improved (compare to Fig.~5 in~\citet{martin2017human}).
-The intended simulations indicate that the accuracy of genetic risk scores is still
+Simulations under the correct model indicate that the accuracy of genetic risk scores is still
 substantially reduced in understudied populations (Fig.~\ref{fig:prs}E-G),
@@
 across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
-At least for the neutral polygenic architectures considered here, genetic drift alone
-does not appear to induce large directional biases in mean predicted risk.
+At least for the neutral polygenic architectures and continental-scale demographic model
+considered here, genetic drift alone does not appear to induce large directional biases
+in mean predicted risk.
 
@@
   population_configurations=population_configurations,
-  demographic_events=demographic_events, # Missing in Martin et al. 2017!
+  demographic_events=demographic_events, # Missing in Martin et al. 2017
   migration_matrix=migration_matrix)
@@
     \textbf{(C)} Marginal allele frequency spectra under the two models. Heterozygosity in the incorrect model
-    is inflated by $~3.5\%$, though the general shape of the distributions are qualitatively similar.
+    is inflated by $\sim3.5\%$, though the general shape of the distributions are qualitatively similar.
     \textbf{(D)} Similarly, the increased heterozygosity leads to excess $D^2$, though the LD-decay is
     qualitatively similar between models.
-    \textbf{(E-G)} True size history for each population plotted against the expected size history from
-    the expected inverse coalescence rates.
+    \textbf{(E-G)} True size history for each population plotted against the expected size history computed
+    from the inverse coalescence rates under each model.
 }
@@
     events in the past, so continental populations were simulated as isolated with low levels of migration
-    for all time. The intended model is shown in Figure \ref{fig:ooa_stats}A.
+    for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
     \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
     demographic model.
-    While risk prediction in the African and East Asian population is reduced compared to the studied
-    European population, the reduction in prediction accuracy is not as hopeless as originally reported.
-    Notably, under a neutral polygenic architecture, we do not expect significant bias in inferred PRS
-    in any of the populations (C, D).
-    Correlations were computed over 100 simulation replicates.
+    GWAS summary statistics were computed from 10,000 case and control subjects in the European
+    population, and the distribution of inferred PRS were compared across the three simulated populations.
+    Unlike the original study, we do not observe large differences in mean inferred PRS across
+    the three populations (C, D), and while risk prediction in the African and East Asian population is
+    still reduced compared to the European population, the reduction in prediction accuracy is not as
+    large as reported in the original study (E-G).
+    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
     For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.

### 2020-05-22 paper.tex
Fix typo

@@
     Unlike the original study, we do not observe large differences in mean inferred PRS across
-    the three populations (C, D), and while risk prediction in the African and East Asian population is
+    the three populations (C, D), and while risk prediction in the African and East Asian populations is
     still reduced compared to the European population, the reduction in prediction accuracy is not as

### 2020-05-22 paper.tex
be nicer to Martin et al

@@
 In particular, we do not observe large differences in mean predicted risk 
-across populations (Fig.~\ref{fig:prs}C,D), as presented in~\citet{martin2017human}.
-At least for the neutral polygenic architectures and continental-scale demographic model
-considered here, genetic drift alone does not appear to induce large directional biases
-in mean predicted risk.
+across populations (Fig.~\ref{fig:prs}C,D), present in data and simulations from~\citet{martin2017human}.
+In other words, a model with continental-scale demographic structure, highly polygenic architecture, and
+neutral evolution does not appear to explain the large directional biases
+in mean predicted risk identified in~\citet{martin2017human}.
 
-This error was easy to make, and could have been prevented by better API design.
+From a software perspective, this error was easy to make, and could have been prevented by better API design.
 The original \msprime\ API requires the user to pass

### 2020-05-22 paper.tex
Minor fiddling

--- 2020-05-22 paper.tex
@@
 in mean predicted risk.
+
+\aprcomment{
+In particular, we do not observe large differences in mean predicted risk 
+across populations (Fig.~\ref{fig:prs}C,D), present in data and simulations from~\citet{martin2017human}.
+Thus, a model with continental-scale demographic structure, highly polygenic architecture, and
+neutral evolution does not appear to explain the observed large directional biases
+in mean predicted risk.
+}
 

### 2020-05-22 paper.tex

--- 2020-05-22 paper.tex
@@
 \documentclass{article}
+\usepackage{fullpage}
 \usepackage[round]{natbib}
 
-\usepackage{fullpage}
 \usepackage{listings}
@@
 across populations (Fig.~\ref{fig:prs}C,D), present in data and simulations from~\citet{martin2017human}.
-In other words, a model with continental-scale demographic structure, highly polygenic architecture, and
+Thus, a model with continental-scale demographic structure, highly polygenic architecture, and
 neutral evolution does not appear to explain the large directional biases
@@
 
-From a software perspective, this error was easy to make, and could have been prevented by better API design.
+From a software perspective, this error was easy to make,
+and could have been prevented by better API design.
 The original \msprime\ API requires the user to pass

### 2020-05-27 analysis/prs_plot.py
Add comparison to Martin et al violins in figure 2

@@ -38,13 +38,14 @@ def adjacent_values(vals, q1, q3):
     return lower_adjacent_value, upper_adjacent_value
 
 
-def set_axis_style(ax, labels):
+def set_axis_style(ax, labels, label_positions):
     ax.get_xaxis().set_tick_params(direction='out')
     ax.xaxis.set_ticks_position('bottom')
-    ax.set_xticks(np.arange(1, len(labels) + 1))
+    ax.set_xticks(label_positions)
     ax.set_xticklabels(labels)
-    ax.set_xlim(0.25, len(labels) + 0.75)
-    ax.set_xlabel('Sample name')
+    ax.set_xlim(label_positions[0]-.75, label_positions[-1]+.75)
+    #ax.set_xlabel('Sample name')
+    ax.set_xlabel('')
 
 
 if __name__ == "__main__":
@@ -143,35 +144,55 @@ if __name__ == "__main__":
     ax4.legend(fontsize=6, frameon=False)
     
     ## violin plots
+    
+    # data from Martin et al for h2 = 0.67:
+    data_martin = pandas.read_csv('data/h2_67_newtree.tsv', sep='\t')
+
+    corrs_orig = {}
+    for m in [200, 500, 1000]:
+        corrs_orig[m] = {}
+        for pop in ['AFR', 'EUR', 'EAS']:
+            corrs_orig[m][pop] = list(data_martin[(data_martin['m']==m) & (data_martin['Pop']==pop)]['estimate'])
+
+    
     h2 = 0.67
-    violin_colors = [cb_colors[0], cb_colors[3], cb_colors[1]]
+    violin_colors = [cb_colors[0], cb_colors[0],
+                     cb_colors[3], cb_colors[3],
+                     cb_colors[1], cb_colors[1]]
+
+    violin_positions = [1.2, 1.8, 3.2, 3.8, 5.2, 5.8]
 
     ax5 = plt.subplot2grid((4, 9), (2, 0), rowspan=2, colspan=3)
-    m = 200
-    data = [correlations[h2][m][k] for k in ['AFR','EAS','EUR']]
     
+    m = 200
+    data = [corrs_orig[m]['AFR'], correlations[h2][m]['AFR'],
+            corrs_orig[m]['EAS'], correlations[h2][m]['EAS'],
+            corrs_orig[m]['EUR'], correlations[h2][m]['EUR']]
+
     for y in [0.25, 0.5, 0.75]:
[…]

### 2020-05-29 paper.tex
Small clarification in Fig 2 caption.
Add comparison to Martin et al violins in figure 2
Moved the ms model and into intro.

@@
 \documentclass{article}
+\usepackage[round]{natbib}
 \usepackage{fullpage}
-\usepackage[round]{natbib}
-
-\usepackage{listings}
-\usepackage{url}
-
 \usepackage{authblk}
 \usepackage{graphicx}
-\usepackage{color}
-
-\lstset{language=Python}
+\usepackage{minted}
+\usepackage{tcolorbox}
+\usepackage{etoolbox}
+\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
+\AfterEndEnvironment{minted}{\end{tcolorbox}}%
+\usepackage{url}
 
@@
 line options to describe demographic models.
-Particularly for the large models we use today, these are not
-intuitive or comprehensible for humans.
-The Python API for \msprime, by contrast, allows the user to
-state models in a more human-readable manner.
-An example contrasting the different approaches to model specification
-is provided in the Methods section.
-%\sgcomment{Other software, such as SLiM,
-%have developed advanced user interfaces and even dedicated programming languages
-%to facilitate the specification of various modes of evolution.\cite{}
-%}
+For example, the~\citet{gutenkunst2009inferring} demographic model
+(that is the subject of this note), as written
+in \ms\ syntax, is
+\begin{minted}[]{tcsh}
+-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
+-eg 0 2 116.010723 -eg 0 3 160.246047
+-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
+-ej 0.028985 3 2 -en 0.028985 2 0.287184
+-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
+-ej 0.197963 2 1 -en 0.303501 1 1
+\end{minted}
+\noindent This model is relatively simple,
+and models with many more populations and parameters are increasingly
+common. Descriptions of such models in \ms\ syntax are not easy to comprehend.
+The Python interface for \msprime, by contrast, allows the user to
+state models in a more human-readable and programmatic manner,
+and has many advantages over \ms's command line interface.
 
-Implementing multi-population models of demographic history is difficult and error-prone.
+Even when using a high-level programming language like Python, however,
+implementing multi-population models
+of demographic history is difficult and error-prone.
 In this note we discuss two implementation errors that arose through
@@
 \msprime\ input.
-% TODO ask Franz about this
 Finally, \cite{pfaffelhuber2020choose}
@@
 between those populations at the same time.
+The demographic events for the correct model are given as
+
+\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
+demographic_events = [
+    # CEU and CHB merge into B with rate changes at T_EU_AS
+    msprime.MassMigration(
+        time=T_EU_AS, source=2, destination=1, proportion=1.0),
+    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
+    msprime.MigrationRateChange(
+        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
+    msprime.MigrationRateChange(
+        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
+    msprime.PopulationParametersChange(
+        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
+    # Population B merges into YRI at T_B
+    msprime.MassMigration(
+        time=T_B, source=1, destination=0, proportion=1.0),
+    msprime.MigrationRateChange(time=T_B, rate=0), # NB This line was missing!!!
+    # Size changes to N_A at T_AF
+    msprime.PopulationParametersChange(
+        time=T_AF, initial_size=N_A, population_id=0)
+]
+\end{minted}
+Note that the migration rate change on 15 was missing in the incorrect model.
 The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
@@
 
-From a software perspective, this error was easy to make,
+From a software perspective, this error was easy to make
 and could have been prevented by better API design.
@@
 three separate parameters to specify a demographic model. For example,
-\begin{lstlisting}[frame=single]
+\begin{minted}[highlightlines={8}]{python}
 dbg = msprime.DemographyDebugger(
@@
   migration_matrix=migration_matrix)
-\end{lstlisting}
+\end{minted}
 Note that the same three parameters must be passed to both the debugger
@@
 snippet would then be rewritten as
-\begin{lstlisting}[frame=single]
+\begin{minted}[]{python}
 demography = msprime.Demography(
@@
 ts = msprime.simulate(demography=demography)
-\end{lstlisting}
+\end{minted}
 This simple change to the interface makes it much less likely
@@
 
-For reference, the~\citet{gutenkunst2009inferring} demographic model, as written in \ms, is
-\begin{lstlisting}[frame=single]
--n 1 1.682020 -n 2 3.736830 -n 3 7.292050
--eg 0 2 116.010723 -eg 0 3 160.246047
--ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
--ej 0.028985 3 2 -en 0.028985 2 0.287184
--ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
--ej 0.197963 2 1 -en 0.303501 1 1
-\end{lstlisting}
-The equivalent model using \msprime\ is more human readable and provided in full at
-\url{https://github.com/jeromekelleher/msprime-model-errors}. The demographic events
-are given as
-\begin{lstlisting}[frame=single]
-demographic_events = [
-    # CEU and CHB merge into B with rate changes at T_EU_AS
-    msprime.MassMigration(
-        time=T_EU_AS, source=2, destination=1, proportion=1.0),
-    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
-    msprime.MigrationRateChange(
-        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
-    msprime.MigrationRateChange(
-        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
-    msprime.PopulationParametersChange(
-        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
-    # Population B merges into YRI at T_B
-    msprime.MassMigration(
-        time=T_B, source=1, destination=0, proportion=1.0),
-    msprime.MigrationRateChange(time=T_B, rate=0),
-    # Size changes to N_A at T_AF
-    msprime.PopulationParametersChange(
-        time=T_AF, initial_size=N_A, population_id=0)
-]
-\end{lstlisting}
-It is worth noting that the~\citet{gutenkunst2009inferring} model is relatively
-simple, and models with many more populations and parameters are increasingly
-common.
+\section*{Acknowledgements}
+\aprcomment{to fill in}
+\begin{itemize}
+\item Alicia - sharing data, comments?
+\item Ilan - comments
+\item Chris for comments?
+\end{itemize}
 
@@
     population, and the distribution of inferred PRS were compared across the three simulated populations.
+    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
     Unlike the original study, we do not observe large differences in mean inferred PRS across
@@
     large as reported in the original study (E-G).
-    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
+    Panel D shows a comparison of PRS distributions for a simulation replicate with 1,000 causal variants.
+    For each population in the violin plots, the original correlations from~\citet{martin2017human}
+    are shown on the left, and correlations using the correct model are shown on the right.
     For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.

### 2020-06-02 paper.tex
Minor tweaks and add acks.

@@
 scientific software developers often focus on computational speed and biological realism.
-Documentation and application programming interface (API), however, also determine
-whether software is used, and whether it is used correctly.
+As the models simulated become more realistic, however, they also become
+more complex and difficult to specify. The interface through which
+users define their models is therefore increasingly important. Without
+an intuitive and thoroughly documented interface, it is very difficult
+to simulate complex population models correctly.
+
+% Documentation and , however, also determine
+% whether software is used, and whether it is used correctly.
 
@@
 samples of whole chromosomes for the first time. Another distinct advantage
-of \msprime\ is its Python API, greatly
-increasing the flexibility and ease of use over the standard approach of
+of \msprime\ is its Python application programming interface (API), which greatly
+increases the flexibility and ease of use over the standard approach of
 text-based command line interfaces.
@@
 \section*{Acknowledgements}
-\aprcomment{to fill in}
-\begin{itemize}
-\item Alicia - sharing data, comments?
-\item Ilan - comments
-\item Chris for comments?
-\end{itemize}
+We would like to thank Alicia Martin for comments on the manuscript
+and for providing the simulation data from~\cite{martin2017human}
+used in Fig.~\ref{fig:prs}. We also like to thank
+Chris Gignoux and Ilan Gronau for helpful comments on the manuscript.
+JK would like to apologise
+to those affected by the error in the example Out-of-Africa model and
+by the poor design decisions in \msprime's demography API.
+
+% JK: I left out the bit about thanking Alicia for the data - seems
+% unnecessary?
+% \aprcomment{to fill in}
+% \begin{itemize}
+% \item Alicia - sharing data, comments?
+% \item Ilan - comments
+% \item Chris for comments?
+% \end{itemize}
 

### 2020-06-10 paper.tex
Add note about mutation rates and thank JN.

@@
 The discovery and correction of the demographic models discussed here underscore
-how API design choice can lead to the propagation of mistakes that are difficult
-to notice. We therefore recommend the following steps to ensure more robust simulations.
+how API design choice can lead to the propagation of mistakes that are difficult to notice.
+There are myriad other ways to get models slightly or very wrong. Migration
+rates can be set in the wrong direction, for example, and parameters can be
+rescaled incorrectly when a new mutation rate is used.
+We therefore recommend the following steps to ensure more robust simulations.
 
@@
 used in Fig.~\ref{fig:prs}. We also like to thank
-Chris Gignoux and Ilan Gronau for helpful comments on the manuscript.
+Chris Gignoux, Ilan Gronau and John Novembre for helpful comments on the manuscript.
 JK would like to apologise

### 2020-08-03 paper.tex
some initial changes to text to start addressing reviewer comments

@@
 \documentclass{article}
-\usepackage[round]{natbib}
+\usepackage[super,comma]{natbib}
 \usepackage{fullpage}
@@
 \affil[1]{McGill Genome Centre and Department of Human Genetics,
-McGill University, Montr\'{e}al, Qu\'{e}bec, Canada}
+McGill University, Montr\'{e}al, Qu\'{e}bec, H2X 3C9, Canada}
 \affil[2]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
-University of Oxford, Oxford, United Kingdom}
+University of Oxford, Oxford, \aprcomment{postal code} United Kingdom}
 \affil[$\star$]{Joint senior authors, listed alphabetically}
@@
 Despite these initial checks being made, the simulation
-was performed without passing the parameter defining demographic events,
-so that the three populations
-never merged and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
-leading to a vast overestimate of the divergence across human populations.
+was performed without passing the list of demographic events such as historical
+changes in size or population splits, so that the three populations
+never merged and remained separated with low levels of migration (Fig~\ref{fig:prs}A).
+This leads to a vast overestimate of the divergence across human populations.
 While the correct model predicts a mean $F_{ST}$ of
@@
 
-Firstly, if possible, we recommend avoiding the implementation of
-demographic models entirely by using \stdpopsim~\citep{adrion2019community}.
+Firstly, if possible, we recommend using the \stdpopsim~\citep{adrion2019community}
+library, which removes the need to re-implement demographic models from scratch.
 This is a ``standard library" of quality-controlled
@@
 case, as processing such a huge dataset (600,000 samples) was
-highly challenging. Recent progress has made statistical analysis this scale much
-easier~\citep{ralph2020efficiently}, but the problems would have been
+highly challenging. Recent progress has made statistical analysis scale much
+easier for large sample sizes~\citep{ralph2020efficiently}, but the problems would have been
 apparent if smaller pilot simulations were undertaken and analysed.
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{unsrtnat}
 \bibliography{paper}
@@
     for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
-    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
+    \textbf{(B-G)} We repeated the simulation experiment in~\citet{martin2017human} using the correct
     demographic model.

### 2020-08-03 paper.tex
figure and caption updates

@@
 
-\section*{Methods}
+\section*{Data and Code Availability}
 
@@
     qualitatively similar between models.
-    \textbf{(E-G)} True size history for each population plotted against the expected size history computed
-    from the inverse coalescence rates under each model.
+    \textbf{(E-G)} We compared the true size history for each population under this model to their
+    expected inverse coalescent rates, which are often interpreted as $N_e$. The correct and incorrect
+    models (A and B) are equivalent in the recent history, so the inverse coalescence rates only differ
+    in the more distant past. For multi-population demographic history, the inverse coalescence rates
+    are not expected to match the ``true'' historical size along each branch, as population structure and
+    migration between populations change coalescent probabilities over time.
 }

### 2020-08-03 analysis/ooa_expected_statistics.py

@@ -108,6 +108,11 @@ if __name__ == "__main__":
         true_sizes[int(T_B):, 0]
     ))
 
+    # population colors, to match Martin et al
+    yri_color = palettes.Category10[5][3]
+    ceu_color = palettes.Category10[5][0]
+    chb_color = palettes.Category10[5][4]
+
     fig = plt.figure(1, figsize=(6.5, 6))
     fig.clf()
 
@@ -129,72 +134,72 @@ if __name__ == "__main__":
 
     ax3 = plt.subplot2grid((3, 6), (1, 0), colspan=3)
     ax3.plot(fs.marginalize([1,2]),
-             '-', lw=1, color=cb_colors[0], label='YRI (model A)')
+             '-', lw=1, color=yri_color, label='YRI (model A)')
     ax3.plot(fs_mig.marginalize([1,2]),
-             '--', lw=1, color=cb_colors[0], label='YRI (model B)')
+             '--', lw=1, color=yri_color, label='YRI (model B)')
     
     ax3.plot(fs.marginalize([0,2]),
-             '-', lw=1, color=cb_colors[1], label='CEU (A)')
+             '-', lw=1, color=ceu_color, label='CEU (A)')
     ax3.plot(fs_mig.marginalize([0,2]),
-             '--', lw=1, color=cb_colors[1], label='CEU (B)')
+             '--', lw=1, color=ceu_color, label='CEU (B)')
 
     ax3.set_yscale('log')
     ax3.set_xlabel('Allele frequency')
     ax3.set_ylabel(r'Count (with $\theta=1$)')
-    ax3.legend(frameon=False)
+    ax3.legend(frameon=False, fontsize=6)
 
     ax4 = plt.subplot2grid((3, 6), (1, 3), colspan=3)
-    ax4.plot(1e2*rs, D2_yri, '-', color=cb_colors[0], lw=1, label='YRI (A)')
-    ax4.plot(1e2*rs, D2_mig_yri, '--', color=cb_colors[0], lw=1, label='YRI (B)')
-    ax4.plot(1e2*rs, D2_chb, '-', color=cb_colors[3], lw=1, label='CHB (A)')
-    ax4.plot(1e2*rs, D2_mig_chb, '--', color=cb_colors[3], lw=1, label='CHB (B)')
+    ax4.plot(1e2*rs, D2_yri, '-', color=yri_color, lw=1, label='YRI (A)')
[…]

### 2020-08-03 analysis/prs_plot.py

@@ -80,6 +80,10 @@ if __name__ == "__main__":
     correlations = pickle.load(open('data/correlations_alpha_0.0.bp', 'rb'))
     distributions = pickle.load(open('data/distributions_alpha_0.0.bp', 'rb'))
     
+    # population colors, to match Martin et al
+    yri_color = palettes.Category10[5][3]
+    ceu_color = palettes.Category10[5][0]
+    chb_color = palettes.Category10[5][4]
     
     fig = plt.figure(2, figsize=(6.5, 4))
     fig.clf()
@@ -93,51 +97,51 @@ if __name__ == "__main__":
     
     ## plot distributions over 100 replicates
     ax2 = plt.subplot2grid((4, 9), (0, 4), colspan=2)
-    sns.distplot(distributions['true'][.67][1000]['EUR']['mean'],
-                 color=cb_colors[1], hist = False, kde = True,
+    sns.distplot(distributions['true'][.67][1000]['AFR']['mean'],
+                 color=yri_color, hist = False, kde = True,
                  kde_kws = {'shade': True, 'linewidth': 1},
                  ax=ax2)
     sns.distplot(distributions['true'][.67][1000]['EAS']['mean'],
-                 color=cb_colors[3], hist = False, kde = True,
+                 color=chb_color, hist = False, kde = True,
                  kde_kws = {'shade': True, 'linewidth': 1},
                  ax=ax2)
-    sns.distplot(distributions['true'][.67][1000]['AFR']['mean'],
-                 color=cb_colors[0], hist = False, kde = True,
+    sns.distplot(distributions['true'][.67][1000]['EUR']['mean'],
+                 color=ceu_color, hist = False, kde = True,
                  kde_kws = {'shade': True, 'linewidth': 1},
                  ax=ax2)
-    
+
     ax2.set_ylabel('Density')
     ax2.set_xlabel('True PRS')
     ax2.set_yticks([0, 0.05])
     ax2.set_xticks([-15, 0, 15])
     
     ax3 = plt.subplot2grid((4, 9), (1, 4), colspan=2)
-    sns.distplot(distributions['infer'][.67][1000]['EUR']['mean'],
-                 color=cb_colors[1], hist = False, kde = True,
+    sns.distplot(distributions['infer'][.67][1000]['AFR']['mean'],
+                 color=yri_color, hist = False, kde = True,
[…]

### 2020-08-05 paper.tex
more edits to figure captions and text

@@
 In most cases the publications used simulations to test a non-demographic inference
-method, and the model was used as an example of a complex population
-history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
+method, such as inferring gene genealogies from sequencing data~\citep{kelleher2019inferring},
+determining the age of a genetic variant~\citep{albers2020dating}, 
+or studying the power of variant association tests~\citep{tong2020population},
+and the model was used as an example of a complex population history.
 \cite{zhou2018popdemog} used the incorrect model as an example
@@
 the misspecified model in \msprime's documentation.
-If the required model is not present in \stdpopsim, then particular care should
+However, \stdpopsim has QC-verified models available for only a handful of published
+models and species, so it might not meet the needs of a particular simulation study.
+If the required model is not present in \stdpopsim, then extra care should
 be taken to validate the implementation.
@@
 highly challenging. Recent progress has made statistical analysis scale much
-easier for large sample sizes~\citep{ralph2020efficiently}, but the problems would have been
-apparent if smaller pilot simulations were undertaken and analysed.
+easier for large sample sizes~\citep{ralph2020efficiently}, so that $F_{ST}$,
+heterozygosity, the allele frequency spectrum, or the distribution of coalescence
+times can be rapidly computed. However, the problems would have been
+apparent if pilot simulations with smaller sample sizes were undertaken and analysed,
+as simulations and analysis of small samples sizes is very efficient.
 Graphical inspection of demographic models can also help identify issues, especially if
@@
 the output of \msprime\ simulations using \tskit~ver.~0.2.3~\citep{ralph2020efficiently}.
-Demographic models were plotted using the \texttt{demography} package
-(\url{https://github.com/apragsdale/demography}, ver.~0.0.3).
-We used the original pipeline from~\citet{martin2017human} available from
+Demographic models were plotted using the \texttt{demography} package written for Python
+(\url{https://github.com/apragsdale/demography}, ver.~0.0.3). The \texttt{demography} package
+allows users to define a demographic model as a directed acyclic graph using
+\texttt{networkx}~\citep{hagberg2008exploring} which can be plotted as in Figures 1A,B and 2A.
+\texttt{demography} also can also translate the graph-based demography into \texttt{msprime} input
+commands or simulate the allele frequency spectrum or two-locus statistics using 
+\texttt{moments}~\citep{jouganous2017inferring, ragsdale2019models}
+or \texttt{dadi}~\citep{gutenkunst2009inferring}.
+
+For the analysis of polygenic risk, we used the original pipeline
+from~\citet{martin2017human} available from
 \url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py}.
@@
     large as reported in the original study (E-G).
+    Panels B and C show the distribution of true and inferred PRS after standardizing PRS across all individuals
+    in the simulation (this is in contrast to Figure 5 in Martin et al., which plotted unstandardized PRS). Because
+    the polygenic trait was simulated as a threshold trait, it is the relative distribution of PRS across indivuduals
+    that determined cases and controls instead of absolute true or inferred PRS.
     Panel D shows a comparison of PRS distributions for a simulation replicate with 1,000 causal variants.

### 2020-08-05 analysis/ooa_expected_statistics.py

@@ -109,9 +109,9 @@ if __name__ == "__main__":
     ))
 
     # population colors, to match Martin et al
-    yri_color = palettes.Category10[5][3]
-    ceu_color = palettes.Category10[5][0]
-    chb_color = palettes.Category10[5][4]
+    yri_color = palettes.Paired[10][5]
+    ceu_color = palettes.Paired[10][1]
+    chb_color = palettes.Paired[10][9]
 
     fig = plt.figure(1, figsize=(6.5, 6))
     fig.clf()
@@ -208,9 +208,9 @@ if __name__ == "__main__":
     fig.text(0.5, 0.98, 'B', fontsize=9, ha='center', va='center')
     fig.text(0.025, 0.65, 'C', fontsize=9, ha='center', va='center')
     fig.text(0.525, 0.65, 'D', fontsize=9, ha='center', va='center')
-    fig.text(0.025, 0.315, 'E', fontsize=9, ha='center', va='center')
-    fig.text(0.35, 0.315, 'F', fontsize=9, ha='center', va='center')
-    fig.text(0.675, 0.315, 'G', fontsize=9, ha='center', va='center')
+    fig.text(0.025, 0.32, 'E', fontsize=9, ha='center', va='center')
+    fig.text(0.35, 0.32, 'F', fontsize=9, ha='center', va='center')
+    fig.text(0.675, 0.32, 'G', fontsize=9, ha='center', va='center')
 
     #fig.tight_layout()
     fig.subplots_adjust(wspace=1.5, hspace=0.4, right=0.975, left=0.1, top=0.975, bottom=.08)

### 2020-08-05 analysis/prs_plot.py

@@ -81,9 +81,9 @@ if __name__ == "__main__":
     distributions = pickle.load(open('data/distributions_alpha_0.0.bp', 'rb'))
     
     # population colors, to match Martin et al
-    yri_color = palettes.Category10[5][3]
-    ceu_color = palettes.Category10[5][0]
-    chb_color = palettes.Category10[5][4]
+    yri_color = palettes.Paired[10][5]
+    ceu_color = palettes.Paired[10][1]
+    chb_color = palettes.Paired[10][9]
     
     fig = plt.figure(2, figsize=(6.5, 4))
     fig.clf()

### 2020-08-05 paper.tex
remove shading around code blocks

@@
 \usepackage{url}
+\usepackage{fancyvrb}
 
@@
 in \ms\ syntax, is
-\begin{minted}[]{tcsh}
+%\begin{minted}[]{tcsh}
+\begin{indent}
+\begin{Verbatim}[xleftmargin=.5in]
 -n 1 1.682020 -n 2 3.736830 -n 3 7.292050
@@
 -ej 0.197963 2 1 -en 0.303501 1 1
-\end{minted}
+\end{Verbatim}
+\end{indent}
+%\end{minted}
 \noindent This model is relatively simple,
@@
 
-\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
+%\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
+\begin{Verbatim}[xleftmargin=.5in]
 demographic_events = [
@@
 ]
-\end{minted}
+\end{Verbatim}
+%\end{minted}
 Note that the migration rate change on 15 was missing in the incorrect model.
@@
 three separate parameters to specify a demographic model. For example,
-\begin{minted}[highlightlines={8}]{python}
+%\begin{minted}[highlightlines={8}]{python}
+\begin{Verbatim}[xleftmargin=.5in]
 dbg = msprime.DemographyDebugger(
@@
   migration_matrix=migration_matrix)
-\end{minted}
+\end{Verbatim}
+%\end{minted}
 Note that the same three parameters must be passed to both the debugger
@@
 snippet would then be rewritten as
-\begin{minted}[]{python}
+%\begin{minted}[]{python}
+\begin{Verbatim}[xleftmargin=.5in]
 demography = msprime.Demography(
@@
 ts = msprime.simulate(demography=demography)
-\end{minted}
+\end{Verbatim}
+%\end{minted}
 This simple change to the interface makes it much less likely
@@
 We would like to thank Alicia Martin for comments on the manuscript
-and for providing the simulation data from~\cite{martin2017human}
-used in Fig.~\ref{fig:prs}. We also like to thank
+and for providing the simulation data from their study reproduced
+in Fig.~\ref{fig:prs}. We also like to thank
 Chris Gignoux, Ilan Gronau and John Novembre for helpful comments on the manuscript.

### 2020-08-05 paper.tex
Minor tweaks.

@@
 \affil[2]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
-University of Oxford, Oxford, \aprcomment{postal code} United Kingdom}
+University of Oxford, Oxford, OX3 7LF, United Kingdom}
 \affil[$\star$]{Joint senior authors, listed alphabetically}
@@
 and the model was used as an example of a complex population history.
-\cite{zhou2018popdemog} used the incorrect model as an example
+\citet{zhou2018popdemog}~used the incorrect model as an example
 of how their method for visualising demographic models can support
 \msprime\ input.
-Finally, \cite{pfaffelhuber2020choose}
-used simulations of the incorrect model demography to evaluate
+Finally, \citet{pfaffelhuber2020choose}~used simulations
+of the incorrect model demography to evaluate
 their method for choosing ancestry informative markers.
@@
 %\end{minted}
-Note that the migration rate change on 15 was missing in the incorrect model.
+Note that the migration rate change on line 15 was missing in the incorrect model.
 The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
@@
 \msprime~ver.~1.0 will introduce a Demography class. The above
-snippet would then be rewritten as
+snippet would then be rewritten (approximately) as
 %\begin{minted}[]{python}
@@
 demography = msprime.Demography(
-  population_configurations=population_configurations,
-  demographic_events=demographic_events,
+  populations=populations,
+  events=demographic_events,
   migration_matrix=migration_matrix)
@@
 dbg.print_history()
-ts = msprime.simulate(demography=demography)
+ts = msprime.simulate(
+  # other parameters defining samples etc
+  demography=demography)
 \end{Verbatim}
@@
 the misspecified model in \msprime's documentation.
-However, \stdpopsim has QC-verified models available for only a handful of published
+However, \stdpopsim\ has QC-verified models available for only a handful of published
 models and species, so it might not meet the needs of a particular simulation study.
@@
 case, as processing such a huge dataset (600,000 samples) was
-highly challenging. Recent progress has made statistical analysis scale much
-easier for large sample sizes~\citep{ralph2020efficiently}, so that $F_{ST}$,
+highly challenging. Recent progress has made statistical analysis at this scale much
+easier~\citep{ralph2020efficiently}, so that $F_{ST}$,
 heterozygosity, the allele frequency spectrum, or the distribution of coalescence

### 2020-08-06 paper.tex
minor tweak for clarity

@@
 determining the age of a genetic variant~\citep{albers2020dating},
-or studying the power of variant association tests~\citep{tong2020population},
-and the model was used as an example of a complex population history.
+or studying the power of variant association tests~\citep{tong2020population}.
+In each of these studies, this model was used as an example of a complex population history.
 \citet{zhou2018popdemog}~used the incorrect model as an example

### 2020-08-19 paper.tex
style edit

--- 2020-08-05 paper.tex
@@
 \documentclass{article}
-\usepackage[super,comma]{natbib}
+\usepackage[round]{natbib}
+
 \usepackage{fullpage}
+\usepackage{listings}
+\usepackage{url}
+
 \usepackage{authblk}
 \usepackage{graphicx}
-\usepackage{minted}
-\usepackage{tcolorbox}
-\usepackage{etoolbox}
-\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
-\AfterEndEnvironment{minted}{\end{tcolorbox}}%
-\usepackage{url}
-\usepackage{fancyvrb}
+\usepackage{color}
+
+\lstset{language=Python}
 
@@
 \affil[1]{McGill Genome Centre and Department of Human Genetics,
-McGill University, Montr\'{e}al, Qu\'{e}bec, H2X 3C9, Canada}
+McGill University, Montr\'{e}al, Qu\'{e}bec, Canada}
 \affil[2]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
-University of Oxford, Oxford, \aprcomment{postal code} United Kingdom}
+University of Oxford, Oxford, United Kingdom}
 \affil[$\star$]{Joint senior authors, listed alphabetically}
@@
 scientific software developers often focus on computational speed and biological realism.
-As the models simulated become more realistic, however, they also become
-more complex and difficult to specify. The interface through which
-users define their models is therefore increasingly important. Without
-an intuitive and thoroughly documented interface, it is very difficult
-to simulate complex population models correctly.
-
-% Documentation and , however, also determine
-% whether software is used, and whether it is used correctly.
+Documentation and application programming interface (API), however, also determine
+whether software is used, and whether it is used correctly.
 
@@
 samples of whole chromosomes for the first time. Another distinct advantage
-of \msprime\ is its Python application programming interface (API), which greatly
-increases the flexibility and ease of use over the standard approach of
+of \msprime\ is its Python API, greatly
+increasing the flexibility and ease of use over the standard approach of
 text-based command line interfaces.
@@
 line options to describe demographic models.
-For example, the~\citet{gutenkunst2009inferring} demographic model
-(that is the subject of this note), as written
-in \ms\ syntax, is
-%\begin{minted}[]{tcsh}
-\begin{indent}
-\begin{Verbatim}[xleftmargin=.5in]
--n 1 1.682020 -n 2 3.736830 -n 3 7.292050
--eg 0 2 116.010723 -eg 0 3 160.246047
--ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
--ej 0.028985 3 2 -en 0.028985 2 0.287184
--ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
--ej 0.197963 2 1 -en 0.303501 1 1
-\end{Verbatim}
-\end{indent}
-%\end{minted}
-\noindent This model is relatively simple,
-and models with many more populations and parameters are increasingly
-common. Descriptions of such models in \ms\ syntax are not easy to comprehend.
-The Python interface for \msprime, by contrast, allows the user to
-state models in a more human-readable and programmatic manner,
-and has many advantages over \ms's command line interface.
+Particularly for the large models we use today, these are not
+intuitive or comprehensible for humans.
+The Python API for \msprime, by contrast, allows the user to
+state models in a more human-readable manner.
+An example contrasting the different approaches to model specification
+is provided in the Methods section.
+%\sgcomment{Other software, such as SLiM,
+%have developed advanced user interfaces and even dedicated programming languages
+%to facilitate the specification of various modes of evolution.\cite{}
+%}
 
-Even when using a high-level programming language like Python, however,
-implementing multi-population models
-of demographic history is difficult and error-prone.
+Implementing multi-population models of demographic history is difficult and error-prone.
 In this note we discuss two implementation errors that arose through
@@
 In most cases the publications used simulations to test a non-demographic inference
-method, such as inferring gene genealogies from sequencing data~\citep{kelleher2019inferring},
-determining the age of a genetic variant~\citep{albers2020dating}, 
-or studying the power of variant association tests~\citep{tong2020population},
-and the model was used as an example of a complex population history.
+method, and the model was used as an example of a complex population
+history~\citep{kelleher2019inferring,albers2020dating,tong2020population}.
 \cite{zhou2018popdemog} used the incorrect model as an example
@@
 \msprime\ input.
+% TODO ask Franz about this
 Finally, \cite{pfaffelhuber2020choose}
@@
 between those populations at the same time.
-The demographic events for the correct model are given as
-
-%\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
-\begin{Verbatim}[xleftmargin=.5in]
-demographic_events = [
-    # CEU and CHB merge into B with rate changes at T_EU_AS
-    msprime.MassMigration(
-        time=T_EU_AS, source=2, destination=1, proportion=1.0),
-    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
-    msprime.MigrationRateChange(
-        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
-    msprime.MigrationRateChange(
-        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
-    msprime.PopulationParametersChange(
-        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
-    # Population B merges into YRI at T_B
-    msprime.MassMigration(
-        time=T_B, source=1, destination=0, proportion=1.0),
-    msprime.MigrationRateChange(time=T_B, rate=0), # NB This line was missing!!!
-    # Size changes to N_A at T_AF
-    msprime.PopulationParametersChange(
-        time=T_AF, initial_size=N_A, population_id=0)
-]
-\end{Verbatim}
-%\end{minted}
-Note that the migration rate change on 15 was missing in the incorrect model.
 The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
@@
 Despite these initial checks being made, the simulation
-was performed without passing the list of demographic events such as historical
-changes in size or population splits, so that the three populations
-never merged and remained separated with low levels of migration (Fig~\ref{fig:prs}A).
-This leads to a vast overestimate of the divergence across human populations.
+was performed without passing the parameter defining demographic events,
+so that the three populations
+never merged and remained separated with low levels of migration (Fig~\ref{fig:prs}A),
+leading to a vast overestimate of the divergence across human populations.
 While the correct model predicts a mean $F_{ST}$ of
@@
 
-From a software perspective, this error was easy to make
-and could have been prevented by better API design.
+From a software perspective, this error was easy to make, and could have been prevented by better API design.
 The original \msprime\ API requires the user to pass
 three separate parameters to specify a demographic model. For example,
-%\begin{minted}[highlightlines={8}]{python}
-\begin{Verbatim}[xleftmargin=.5in]
+\begin{lstlisting}[frame=single]
 dbg = msprime.DemographyDebugger(
@@
   migration_matrix=migration_matrix)
-\end{Verbatim}
-%\end{minted}
+\end{lstlisting}
 Note that the same three parameters must be passed to both the debugger
@@
 snippet would then be rewritten as
-%\begin{minted}[]{python}
-\begin{Verbatim}[xleftmargin=.5in]
+\begin{lstlisting}[frame=single]
 demography = msprime.Demography(
@@
 ts = msprime.simulate(demography=demography)
-\end{Verbatim}
-%\end{minted}
+\end{lstlisting}
 This simple change to the interface makes it much less likely
@@
 The discovery and correction of the demographic models discussed here underscore
-how API design choice can lead to the propagation of mistakes that are difficult to notice.
-There are myriad other ways to get models slightly or very wrong. Migration
-rates can be set in the wrong direction, for example, and parameters can be
-rescaled incorrectly when a new mutation rate is used.
-We therefore recommend the following steps to ensure more robust simulations.
+how API design choice can lead to the propagation of mistakes that are difficult
+to notice. We therefore recommend the following steps to ensure more robust simulations.
 
-Firstly, if possible, we recommend using the \stdpopsim~\citep{adrion2019community}
-library, which removes the need to re-implement demographic models from scratch.
+Firstly, if possible, we recommend avoiding the implementation of
+demographic models entirely by using \stdpopsim~\citep{adrion2019community}.
 This is a ``standard library" of quality-controlled
@@
 the misspecified model in \msprime's documentation.
-However, \stdpopsim has QC-verified models available for only a handful of published
-models and species, so it might not meet the needs of a particular simulation study.
-If the required model is not present in \stdpopsim, then extra care should
+If the required model is not present in \stdpopsim, then particular care should
 be taken to validate the implementation.
@@
 case, as processing such a huge dataset (600,000 samples) was
-highly challenging. Recent progress has made statistical analysis scale much
-easier for large sample sizes~\citep{ralph2020efficiently}, so that $F_{ST}$,
-heterozygosity, the allele frequency spectrum, or the distribution of coalescence
-times can be rapidly computed. However, the problems would have been
-apparent if pilot simulations with smaller sample sizes were undertaken and analysed,
-as simulations and analysis of small samples sizes is very efficient.
+highly challenging. Recent progress has made statistical analysis this scale much
+easier~\citep{ralph2020efficiently}, but the problems would have been
+apparent if smaller pilot simulations were undertaken and analysed.
 Graphical inspection of demographic models can also help identify issues, especially if
@@
 
-\section*{Data and Code Availability}
+\section*{Methods}
 
@@
 the output of \msprime\ simulations using \tskit~ver.~0.2.3~\citep{ralph2020efficiently}.
-Demographic models were plotted using the \texttt{demography} package written for Python
-(\url{https://github.com/apragsdale/demography}, ver.~0.0.3). The \texttt{demography} package
-allows users to define a demographic model as a directed acyclic graph using
-\texttt{networkx}~\citep{hagberg2008exploring} which can be plotted as in Figures 1A,B and 2A.
-\texttt{demography} also can also translate the graph-based demography into \texttt{msprime} input
-commands or simulate the allele frequency spectrum or two-locus statistics using 
-\texttt{moments}~\citep{jouganous2017inferring, ragsdale2019models}
-or \texttt{dadi}~\citep{gutenkunst2009inferring}.
-
-For the analysis of polygenic risk, we used the original pipeline
-from~\citet{martin2017human} available from
+Demographic models were plotted using the \texttt{demography} package
+(\url{https://github.com/apragsdale/demography}, ver.~0.0.3).
+We used the original pipeline from~\citet{martin2017human} available from
 \url{https://github.com/armartin/ancestry_pipeline/blob/master/simulate_prs.py}.
@@
 
-\section*{Acknowledgements}
-We would like to thank Alicia Martin for comments on the manuscript
-and for providing the simulation data from their study reproduced
-in Fig.~\ref{fig:prs}. We also like to thank
-Chris Gignoux, Ilan Gronau and John Novembre for helpful comments on the manuscript.
-JK would like to apologise
-to those affected by the error in the example Out-of-Africa model and
-by the poor design decisions in \msprime's demography API.
+For reference, the~\citet{gutenkunst2009inferring} demographic model, as written in \ms, is
+\begin{lstlisting}[frame=single]
+-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
+-eg 0 2 116.010723 -eg 0 3 160.246047
+-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
+-ej 0.028985 3 2 -en 0.028985 2 0.287184
+-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
+-ej 0.197963 2 1 -en 0.303501 1 1
+\end{lstlisting}
+The equivalent model using \msprime\ is more human readable and provided in full at
+\url{https://github.com/jeromekelleher/msprime-model-errors}. The demographic events
+are given as
+\begin{lstlisting}[frame=single]
+demographic_events = [
+    # CEU and CHB merge into B with rate changes at T_EU_AS
+    msprime.MassMigration(
+        time=T_EU_AS, source=2, destination=1, proportion=1.0),
+    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
+    msprime.MigrationRateChange(
+        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
+    msprime.MigrationRateChange(
+        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
+    msprime.PopulationParametersChange(
+        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
+    # Population B merges into YRI at T_B
+    msprime.MassMigration(
+        time=T_B, source=1, destination=0, proportion=1.0),
+    msprime.MigrationRateChange(time=T_B, rate=0),
+    # Size changes to N_A at T_AF
+    msprime.PopulationParametersChange(
+        time=T_AF, initial_size=N_A, population_id=0)
+]
+\end{lstlisting}
+It is worth noting that the~\citet{gutenkunst2009inferring} model is relatively
+simple, and models with many more populations and parameters are increasingly
+common.
 
-% JK: I left out the bit about thanking Alicia for the data - seems
-% unnecessary?
-% \aprcomment{to fill in}
-% \begin{itemize}
-% \item Alicia - sharing data, comments?
-% \item Ilan - comments
-% \item Chris for comments?
-% \end{itemize}
-
-\bibliographystyle{unsrtnat}
+\bibliographystyle{plainnat}
 \bibliography{paper}
@@
     qualitatively similar between models.
-    \textbf{(E-G)} We compared the true size history for each population under this model to their
-    expected inverse coalescent rates, which are often interpreted as $N_e$. The correct and incorrect
-    models (A and B) are equivalent in the recent history, so the inverse coalescence rates only differ
-    in the more distant past. For multi-population demographic history, the inverse coalescence rates
-    are not expected to match the ``true'' historical size along each branch, as population structure and
-    migration between populations change coalescent probabilities over time.
+    \textbf{(E-G)} True size history for each population plotted against the expected size history computed
+    from the inverse coalescence rates under each model.
 }
@@
     for all time. The correct model has historical events as shown in Fig.~\ref{fig:ooa_stats}A.
-    \textbf{(B-G)} We repeated the simulation experiment in~\citet{martin2017human} using the correct
+    \textbf{(B-G)} We repeated the simulation experiment in~\cite{martin2017human} using the correct
     demographic model.
@@
     population, and the distribution of inferred PRS were compared across the three simulated populations.
-    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
     Unlike the original study, we do not observe large differences in mean inferred PRS across
@@
     large as reported in the original study (E-G).
-    Panels B and C show the distribution of true and inferred PRS after standardizing PRS across all individuals
-    in the simulation (this is in contrast to Figure 5 in Martin et al., which plotted unstandardized PRS). Because
-    the polygenic trait was simulated as a threshold trait, it is the relative distribution of PRS across indivuduals
-    that determined cases and controls instead of absolute true or inferred PRS.
-    Panel D shows a comparison of PRS distributions for a simulation replicate with 1,000 causal variants.
-    For each population in the violin plots, the original correlations from~\citet{martin2017human}
-    are shown on the left, and correlations using the correct model are shown on the right.
+    Correlations between true and inferred polygenic scores were computed over 100 simulation replicates.
     For direct comparison to the original study, see Figure~5 in~\citet{martin2017human}.

### 2020-08-19 paper.tex
added line numbers to code where we were referring to it

--- 2020-08-06 paper.tex
@@
 %\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
-\begin{Verbatim}[xleftmargin=.5in]
+\begin{Verbatim}[numbers=left, xleftmargin=.5in]
 demographic_events = [
@@
 due diligence. For example, a brief analysis of $F_{ST}$ values
-would likely would have prevented the error in~\citet{martin2017human}.
+likely would have prevented the error in~\citet{martin2017human}.
 It is understandable that such analyses were not undertaken in this

### 2020-08-19 paper_compilable.tex
created compilable tex document for AJHG

--- 2020-08-19 paper.tex
@@
 \usepackage{graphicx}
-\usepackage{minted}
-\usepackage{tcolorbox}
-\usepackage{etoolbox}
-\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
-\AfterEndEnvironment{minted}{\end{tcolorbox}}%
+%\usepackage{minted}
+%\usepackage{tcolorbox}
+%\usepackage{etoolbox}
+%\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
+%\AfterEndEnvironment{minted}{\end{tcolorbox}}%
 \usepackage{url}
@@
 \bibliographystyle{unsrtnat}
-\bibliography{paper}
+\begin{thebibliography}{18}
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\url}[1]{\texttt{#1}}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[Kelleher et~al.(2016)Kelleher, Etheridge, and
+  McVean]{kelleher2016efficient}
+Jerome Kelleher, Alison~M Etheridge, and Gilean McVean.
+\newblock Efficient coalescent simulation and genealogical analysis for large
+  sample sizes.
+\newblock \emph{PLoS computational biology}, 12\penalty0 (5):\penalty0
+  e1004842, 2016.
+
+\bibitem[Nelson et~al.(2020)Nelson, Kelleher, Ragsdale, Moreau, McVean, and
+  Gravel]{nelson2020accounting}
+Dominic Nelson, Jerome Kelleher, Aaron~P Ragsdale, Claudia Moreau, Gil McVean,
+  and Simon Gravel.
+\newblock Accounting for long-range correlations in genome-wide simulations of
+  large cohorts.
+\newblock \emph{PLoS genetics}, 16\penalty0 (5):\penalty0 e1008619, 2020.
+
+\bibitem[Kelleher and Lohse(2020)]{kelleher2020coalescent}
+Jerome Kelleher and Konrad Lohse.
+\newblock Coalescent simulation with msprime.
+\newblock In Julien~Y. Dutheil, editor, \emph{Statistical Population Genomics},
+  pages 191--230. Springer US, New York, NY, 2020.
+
+\bibitem[Hudson(2002)]{hudson2002generating}
+Richard~R. Hudson.
+\newblock Generating samples under a {Wright-Fisher} neutral model of genetic
+  variation.
+\newblock \emph{Bioinformatics}, 18\penalty0 (2):\penalty0 337--338, 2002.
+
+\bibitem[Gutenkunst et~al.(2009)Gutenkunst, Hernandez, Williamson, and
+  Bustamante]{gutenkunst2009inferring}
+Ryan~N Gutenkunst, Ryan~D Hernandez, Scott~H Williamson, and Carlos~D
+  Bustamante.
+\newblock Inferring the joint demographic history of multiple populations from
+  multidimensional snp frequency data.
+\newblock \emph{PLoS genetics}, 5\penalty0 (10):\penalty0 e1000695, 2009.
+
+\bibitem[Kelleher et~al.(2019)Kelleher, Wong, Wohns, Fadil, Albers, and
+  McVean]{kelleher2019inferring}
+Jerome Kelleher, Yan Wong, Anthony~W Wohns, Chaimaa Fadil, Patrick~K Albers,
+  and Gil McVean.
+\newblock Inferring whole-genome histories in large population datasets.
+\newblock \emph{Nature genetics}, 51\penalty0 (9):\penalty0 1330--1338, 2019.
+
+\bibitem[Albers and McVean(2020)]{albers2020dating}
+Patrick~K Albers and Gil McVean.
+\newblock Dating genomic variants and shared ancestry in population-scale
+  sequencing data.
+\newblock \emph{PLoS biology}, 18\penalty0 (1):\penalty0 e3000586, 2020.
+
+\bibitem[Tong and Hernandez(2020)]{tong2020population}
+Dominic~MH Tong and Ryan~D Hernandez.
+\newblock Population genetic simulation study of power in association testing
+  across genetic architectures and study designs.
+\newblock \emph{Genetic epidemiology}, 44\penalty0 (1):\penalty0 90--103, 2020.
+
+\bibitem[Zhou et~al.(2018)Zhou, Tian, Browning, and Browning]{zhou2018popdemog}
+Ying Zhou, Xiaowen Tian, Brian~L Browning, and Sharon~R Browning.
+\newblock {POPdemog}: visualizing population demographic history from
+  simulation scripts.
+\newblock \emph{Bioinformatics}, 34\penalty0 (16):\penalty0 2854--2855, 2018.
+
+\bibitem[Pfaffelhuber et~al.(2020)Pfaffelhuber, Grundner-Culemann, Lipphardt,
+  and Baumdicker]{pfaffelhuber2020choose}
+Peter Pfaffelhuber, Franziska Grundner-Culemann, Veronika Lipphardt, and Franz
+  Baumdicker.
+\newblock How to choose sets of ancestry informative markers: A supervised
+  feature selection approach.
+\newblock \emph{Forensic Science International: Genetics}, 46:\penalty0 102259,
+  2020.
+
+\bibitem[Martin et~al.(2017)Martin, Gignoux, Walters, Wojcik, Neale, Gravel,
+  Daly, Bustamante, and Kenny]{martin2017human}
+Alicia~R Martin, Christopher~R Gignoux, Raymond~K Walters, Genevieve~L Wojcik,
+  Benjamin~M Neale, Simon Gravel, Mark~J Daly, Carlos~D Bustamante, and
+  Eimear~E Kenny.
+\newblock Human demographic history impacts genetic risk prediction across
+  diverse populations.
+\newblock \emph{The American Journal of Human Genetics}, 100\penalty0
+  (4):\penalty0 635--649, 2017.
+
+\bibitem[Gravel et~al.(2011)Gravel, Henn, Gutenkunst, Indap, Marth, Clark, Yu,
+  Gibbs, Bustamante, Project, et~al.]{gravel2011demographic}
+Simon Gravel, Brenna~M Henn, Ryan~N Gutenkunst, Amit~R Indap, Gabor~T Marth,
+  Andrew~G Clark, Fuli Yu, Richard~A Gibbs, Carlos~D Bustamante, 1000~Genomes
+  Project, et~al.
+\newblock Demographic history and rare allele sharing among human populations.
+\newblock \emph{Proceedings of the National Academy of Sciences}, 108\penalty0
+  (29):\penalty0 11983--11988, 2011.
+
+\bibitem[Popejoy and Fullerton(2016)]{popejoy2016genomics}
+Alice~B Popejoy and Stephanie~M Fullerton.
+\newblock Genomics is failing on diversity.
+\newblock \emph{Nature News}, 538\penalty0 (7624):\penalty0 161, 2016.
+
+\bibitem[Adrion et~al.(2020)Adrion, Cole, Dukler, Galloway, Gladstein, Gower,
+  Kyriazis, Ragsdale, Tsambos, Baumdicker, et~al.]{adrion2019community}
+Jeffrey~R Adrion, Christopher~B Cole, Noah Dukler, Jared~G Galloway, Ariella~L
+  Gladstein, Graham Gower, Christopher~C Kyriazis, Aaron~P Ragsdale, Georgia
+  Tsambos, Franz Baumdicker, et~al.
+\newblock A community-maintained standard library of population genetic models.
+\newblock \emph{eLife}, 9:\penalty0 e54967, 2020.
+
+\bibitem[Ralph et~al.(2020)Ralph, Thornton, and Kelleher]{ralph2020efficiently}
+Peter Ralph, Kevin Thornton, and Jerome Kelleher.
+\newblock Efficiently summarizing relationships in large samples: a general
+  duality between statistics of genealogies and genomes.
+\newblock \emph{Genetics}, 2020.
+
+\bibitem[Jouganous et~al.(2017)Jouganous, Long, Ragsdale, and
+  Gravel]{jouganous2017inferring}
+Julien Jouganous, Will Long, Aaron~P Ragsdale, and Simon Gravel.
+\newblock Inferring the joint demographic history of multiple populations:
+  beyond the diffusion approximation.
+\newblock \emph{Genetics}, 206\penalty0 (3):\penalty0 1549--1567, 2017.
+
+\bibitem[Ragsdale and Gravel(2019)]{ragsdale2019models}
+Aaron~P Ragsdale and Simon Gravel.
+\newblock Models of archaic admixture and recent history from two-locus
+  statistics.
+\newblock \emph{PLoS genetics}, 15\penalty0 (6):\penalty0 e1008204, 2019.
+
+\bibitem[Hagberg et~al.(2008)Hagberg, Swart, and S~Chult]{hagberg2008exploring}
+Aric Hagberg, Pieter Swart, and Daniel S~Chult.
+\newblock Exploring network structure, dynamics, and function using networkx.
+\newblock Technical report, Los Alamos National Lab.(LANL), Los Alamos, NM
+  (United States), 2008.
+
+\end{thebibliography}
 
@@
 \begin{center}
-\makebox[\textwidth][c]{\includegraphics{figures/ooa_expected_stats.pdf}}
+\makebox[\textwidth][c]{\includegraphics{ooa_expected_stats.pdf}}
 \caption{\textbf{Expected diversity statistics under the \citet{gutenkunst2009inferring} model}.
@@
 \begin{center}
-\makebox[\textwidth][c]{\includegraphics{figures/prs_fig.pdf}}
+\makebox[\textwidth][c]{\includegraphics{prs_fig.pdf}}
 \caption{\textbf{The transferability of PRS under neutrality}.

### 2020-09-21 code_blocks/code_blocks.tex
Figure with code blocks

\documentclass{article}
\usepackage{minted}
\usepackage{tcolorbox}
\usepackage{etoolbox}
\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
\AfterEndEnvironment{minted}{\end{tcolorbox}}%
\usepackage{url}

\usepackage[
    paperwidth=4.5in, % 27 pica for two column width in AJHG
    paperheight=7.5in,
    left=0.15in,
    right=0.075in,
    top=0.075in,
    bottom=0.05in
]{geometry}

\begin{document}

\begin{minted}[fontsize=\footnotesize]{tcsh}
-n 1 1.682020 -n 2 3.736830 -n 3 7.292050
-eg 0 2 116.010723 -eg 0 3 160.246047
-ma x 0.881098 0.561966 0.881098 x 2.797460 0.561966 2.797460 x
-ej 0.028985 3 2 -en 0.028985 2 0.287184
-ema 0.028985 3 x 7.293140 x 7.293140 x x x x x
-ej 0.197963 2 1 -en 0.303501 1 1
\end{minted}

\begin{minted}[linenos,highlightlines={16,17},numbersep=5pt,fontsize=\footnotesize]{python}
demographic_events = [
    # CEU and CHB merge into B with rate changes at T_EU_AS
    msprime.MassMigration(
        time=T_EU_AS, source=2, destination=1, proportion=1.0),
    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
    msprime.PopulationParametersChange(
        time=T_EU_AS, initial_size=N_B, growth_rate=0,
        population_id=1),
    # Population B merges into YRI at T_B
    msprime.MassMigration(
        time=T_B, source=1, destination=0, proportion=1.0),
    # Missing in msprime documentation
    msprime.MigrationRateChange(time=T_B, rate=0),
    # Size changes to N_A at T_AF
    msprime.PopulationParametersChange(
        time=T_AF, initial_size=N_A, population_id=0)
]
\end{minted}

\begin{minted}[highlightlines={9},fontsize=\footnotesize]{python}
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg.print_history()
ts = msprime.simulate(
  ..., # parameters defining samples, sequence length, etc
  population_configurations=population_configurations,
  demographic_events=demographic_events, # Missing in Ref. 11
  migration_matrix=migration_matrix)
\end{minted}

\begin{minted}[fontsize=\footnotesize]{python}
demography = msprime.Demography(
  populations=populations,
  events=demographic_events,
  migration_matrix=migration_matrix)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(
  ..., # parameters defining samples, sequence length, etc
  demography=demography)
\end{minted}

\end{document}

### 2020-09-21 analysis/prs_plot.py

@@ -192,7 +192,7 @@ if __name__ == "__main__":
     ax5.scatter(violin_positions, medians, marker=".", color='white', s=8, zorder=3)
     ax5.vlines(violin_positions, quartile1, quartile3, color='black', linestyle='-', lw=3)
     
-    quartile1, medians, quartile3 = np.array([np.percentile(d, [5, 50, 95]) for d in data]).T
+    quartile1, medians, quartile3 = np.array([np.percentile(d, [2.5, 50, 97.5]) for d in data]).T
     ax5.vlines(violin_positions, quartile1, quartile3, color='black', linestyle='-', lw=1)
 
     labels = ['AFR','EAS','EUR']
@@ -230,7 +230,7 @@ if __name__ == "__main__":
     ax6.scatter(violin_positions, medians, marker=".", color='white', s=8, zorder=3)
     ax6.vlines(violin_positions, quartile1, quartile3, color='black', linestyle='-', lw=3)
     
-    quartile1, medians, quartile3 = np.array([np.percentile(d, [5, 50, 95]) for d in data]).T
+    quartile1, medians, quartile3 = np.array([np.percentile(d, [2.5, 50, 97.5]) for d in data]).T
     ax6.vlines(violin_positions, quartile1, quartile3, color='black', linestyle='-', lw=1)
 
     labels = ['AFR','EAS','EUR']
@@ -267,7 +267,7 @@ if __name__ == "__main__":
     ax7.scatter(violin_positions, medians, marker=".", color='white', s=8, zorder=3)
     ax7.vlines(violin_positions, quartile1, quartile3, color='black', linestyle='-', lw=3)
     
-    quartile1, medians, quartile3 = np.array([np.percentile(d, [5, 50, 95]) for d in data]).T
+    quartile1, medians, quartile3 = np.array([np.percentile(d, [2.5, 50, 97.5]) for d in data]).T
     ax7.vlines(violin_positions, quartile1, quartile3, color='black', linestyle='-', lw=1)
 
     labels = ['AFR','EAS','EUR']
@@ -288,5 +288,5 @@ if __name__ == "__main__":
     fig.text(0.685, 0.45, 'G', fontsize=9, ha='center', va='center')
 
     fig.tight_layout()
-    plt.savefig('../figures/prs_fig.pdf', dpi=300)
+    plt.savefig('../prs_fig.pdf', dpi=300)
     plt.show()
